# Package Downloads
```{r message=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager", repos = "https://cloud.r-project.org")
BiocManager::install(version = "3.21")
if (!requireNamespace("DESeq2", quietly = TRUE)) {
  BiocManager::install("DESeq2")
}
install.packages("devtools", repos = "https://cloud.r-project.org")
install.packages("GenomeInfoDb", repos = "https://cloud.r-project.org")
install.packages("tidyverse", repos = "https://cloud.r-project.org")
# Install the Homo Sapiens package
if (!("org.Hs.eg.db" %in% installed.packages())) {
  BiocManager::install("org.Hs.eg.db", update = FALSE)
}

```

# Section 1a

```{r}
# Attach the library
library(org.Hs.eg.db)

# We will need this so we can use the pipe: %>%
library(magrittr)

# Create the data folder if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Define the file path to the plots directory
plots_dir <- "plots"

# Create the plots folder if it doesn't exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}

# Define the file path to the results directory
results_dir <- "results"

# Create the results folder if it doesn't exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Define the file path to the data directory
data_dir <- file.path("data", "SRP192714")

# Declare the file path to the gene expression matrix file
data_file <- file.path(data_dir, "SRP192714.tsv")

# Read in data TSV file
expression_df <- readr::read_tsv(data_file) %>%
  tibble::column_to_rownames("Gene")

# Load refine.bio metadata
refinebio_meta <- readr::read_tsv(file.path(data_dir, "metadata_SRP192714.tsv"))
rownames(refinebio_meta) <- refinebio_meta$refinebio_accession_code

# Load GEO metadata (CSV)
geo_meta <- readr::read_csv("data/GSE129882_PhenoData.transcript.csv")

# Merge on 'refinebio_title' (refine.bio) and 'Sample' (GEO), keep 'refinebio_title' as column name
if ("Sample" %in% colnames(geo_meta) && "refinebio_title" %in% colnames(refinebio_meta)) {
  merged_meta <- dplyr::left_join(refinebio_meta, geo_meta, by = c("refinebio_title" = "Sample"))
} else {
  merged_meta <- refinebio_meta
}

# Create trimmed and full versions of expression and metadata
sample_names <- colnames(expression_df)[colnames(expression_df) != "Gene" & colnames(expression_df) != "Symbol"]
trimmed_samples <- sample_names[1:50]
expression_df_trimmed <- expression_df %>% dplyr::select(all_of(trimmed_samples))
merged_meta_trimmed <- merged_meta %>% dplyr::filter(refinebio_accession_code %in% trimmed_samples)
expression_df_full <- expression_df %>% dplyr::select(all_of(sample_names))
merged_meta_full <- merged_meta

# Set this variable to TRUE to use trimmed data, FALSE for full data
use_trimmed <- TRUE

```

# Section 1b

```{r}
# Bring back the "Gene" column in preparation for mapping
expression_df <- expression_df %>%
  tibble::rownames_to_column("Gene")

# Map Ensembl IDs to their first mapped Symbol
gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys = expression_df$Gene,
  keytype = "ENSEMBL",
  column = "SYMBOL",
  multiVals = "first"
)

# Add the mapped symbols as a new column
expression_df$Symbol <- gene_symbols[expression_df$Gene]

# Reorder columns to have Gene, Symbol, then the rest
expression_df <- expression_df %>%
  dplyr::select(Gene, Symbol, dplyr::everything(), -Gene)

# Write mapped data frame to output file
readr::write_tsv(expression_df, file.path(
  results_dir,
  "SRP192714_Symbols.tsv"
))
```

# Section 1c
```{r}

# Get matrix size
matrix_dim <- dim(expression_df)
cat("Expression matrix dimensions (genes x samples):", matrix_dim[1], "x", matrix_dim[2], "\n")

# Number of genes
cat("Number of genes:", nrow(expression_df), "\n")

# Select only numeric columns for log transformation
expr_numeric <- expression_df %>% dplyr::select(where(is.numeric))

# Log-scale the data (add pseudocount to avoid log(0))
log_expr <- log2(expr_numeric + 1)

# Calculate per-gene median expression
gene_medians <- apply(log_expr, 1, median, na.rm = TRUE)

# Show summary statistics for gene medians
summary(gene_medians)

# Density plot of per-gene median expression
library(ggplot2)
plot_obj <- ggplot(data.frame(median=gene_medians), aes(x=median)) +
  geom_density(fill="skyblue", alpha=0.5) +
  labs(title="Density of Per-Gene Median Expression (log2 scale)",
       x="Median log2(expression + 1)",
       y="Density")

# Save plot to the plots directory
plot_path <- file.path(plots_dir, "per_gene_median_density.png")
ggsave(plot_path, plot=plot_obj, width=6, height=4, dpi=300)

# Also print the plot in the notebook
plot_obj
```

The dataset contains 43,363 genes measured across 1,022 samples. The summary statistics show that most genes have low median expression (median ≈ 0.26 on log2 scale), with a long tail of higher expression values (max ≈ 7.6). From our research, this is typical for transcriptome data, where a small number of genes are highly expressed while the majority have low expression. The density plot visualizes this distribution, showing a peak at lower expression values and a gradual decline towards higher values. Log transformation helps to reduce the impact of extreme values and makes the distribution more interpretable.

# Section 2a-c

```{r}
library(DESeq2)

# --- FULL DATA VERSION (commented out) ---
# # Use numeric columns from expression_df as count matrix
# cts <- expression_df %>% dplyr::select(where(is.numeric)) %>% as.matrix()
# 
# # Update coldata for DESeq2
# coldata <- merged_meta
# rownames(coldata) <- coldata$refinebio_accession_code
# 
# # Subset cts columns to match coldata rownames
# cts <- cts[, rownames(coldata)]
# 
# # Round count matrix to integers for DESeq2
# cts <- round(cts)
# 
# # Check dimensions
# stopifnot(ncol(cts) == nrow(coldata))
# 
# # Choose a grouping column for DESeq2 (update as needed)
# # For example, use 'refinebio_title' or a column from GEO metadata
# dds <- DESeqDataSetFromMatrix(countData = cts,
#                               colData = coldata,
#                               design = ~ refinebio_title)
# 
# dds <- DESeq(dds)
# vsd <- vst(dds, blind=FALSE)
# plotPCA(vsd, intgroup="refinebio_title")

## --- ANALYSIS SECTION: works for trimmed or full data ---
# Choose which dataset to use
if (use_trimmed) {
  expr_mat <- expression_df_trimmed
  meta <- merged_meta_trimmed
  plot_title <- "PCA of First 100 Samples (colored by Exposure)"
  plot_file <- "trimmed_pca_exposure.png"
} else {
  expr_mat <- expression_df_full
  meta <- merged_meta_full
  plot_title <- "PCA of All Samples (colored by Exposure)"
  plot_file <- "full_pca_exposure.png"
}
# Convert to numeric matrix
cts <- as.matrix(sapply(expr_mat, as.numeric))
# Subset metadata to match columns
meta <- meta %>% dplyr::filter(refinebio_accession_code %in% colnames(expr_mat))
rownames(meta) <- meta$refinebio_accession_code
# Ensure columns and rows match
cts <- cts[, rownames(meta)]
# Remove negative values (set to zero)
cts[cts < 0] <- 0
cts <- round(cts)
stopifnot(ncol(cts) == nrow(meta))
# Run DESeq2
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = meta,
                              design = ~ refinebio_title)
dds <- DESeq(dds)
vsd <- vst(dds, blind=FALSE)
# Extract PCA data for custom plot
pca_data <- plotPCA(vsd, intgroup = c("Exposure"), returnData = TRUE)
# Add sample names for merging
pca_data$Sample <- rownames(pca_data)
# Merge Exposure info from metadata
pca_data <- dplyr::left_join(pca_data, meta %>% dplyr::select(refinebio_accession_code, Exposure), by = c("Sample" = "refinebio_accession_code"))
# If Exposure column is missing, use meta$Exposure directly
if (!"Exposure" %in% colnames(pca_data) || all(is.na(pca_data$Exposure))) {
  pca_data$Exposure <- meta$Exposure[match(pca_data$Sample, meta$refinebio_accession_code)]
}
# Plot with ggplot2, color by Exposure
pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = Exposure)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = plot_title, x = "PC1", y = "PC2") +
  theme_minimal()
# Save PCA plot to plots directory
ggsave(filename = file.path(plots_dir, plot_file), plot = pca_plot, width = 6, height = 4, dpi = 300)
# Print plot in notebook
pca_plot
```

# Section 2d i
```{r}

```

# Section 2d ii
```{r}

```

# Section 2e

# Section 2f
```{r}

```

# Section 3a-b
```{r attach library, echo = FALSE}
# Attach the DESeq2 library
library(DESeq2)

# Attach the ggplot2 library for plotting
library(ggplot2)

# We will need this so we can use the pipe: %>%
library(magrittr)

set.seed(12345)
```

```{r Read-in Metadata}
metadata_file <- file.path(data_dir, "metadata_SRP192714.tsv")
metadata <- readr::read_tsv("data/SRP192714/metadata_SRP192714.tsv")
```


```{r Whatever the fuck this is}
# # Make the data in the order of the metadata
# expression_df <- expression_df %>%
#   dplyr::select(metadata$refinebio_accession_code)

# # Check if this is in the same order
# all.equal(colnames(expression_df), metadata$refinebio_accession_code)

# metadata <- metadata %>%
#   # Let's get the RPL10 mutation status from this variable
#   dplyr::mutate(mutation_status = dplyr::case_when(
#     stringr::str_detect(refinebio_title, "R98S") ~ "R98S",
#     stringr::str_detect(refinebio_title, "WT") ~ "reference"
#   ))

# # Print out a preview of `mutation_status`
# str(metadata$mutation_status)

# # Make mutation_status a factor and set the levels appropriately
# metadata <- metadata %>%
#   dplyr::mutate(
#     # Here we define the values our factor variable can have and their order.
#     mutation_status = factor(mutation_status, levels = c("reference", "R98S"))
#   )

# # Define a minimum counts cutoff and filter the data to include
# # only rows (genes) that have total counts above the cutoff
# filtered_expression_df <- expression_df %>%
#   dplyr::filter(rowSums(.) >= 10)

# # round all expression counts
# gene_matrix <- round(filtered_expression_df)

# ddset <- DESeqDataSetFromMatrix(
#   # Here we supply non-normalized count data
#   countData = gene_matrix,
#   # Supply the `colData` with our metadata data frame
#   colData = metadata,
#   # Supply our experimental variable to `design`
#   design = ~mutation_status
# )

# deseq_object <- DESeq(ddset)
# deseq_results <- results(deseq_object)

# deseq_results <- lfcShrink(
#   deseq_object, # The original DESeq2 object after running DESeq()
#   coef = 2, # The log fold change coefficient used in DESeq(); the default is 2.
#   res = deseq_results # The original DESeq2 results table
# )

# # this is of class DESeqResults -- we want a data frame
# deseq_df <- deseq_results %>%
#   # make into data.frame
#   as.data.frame() %>%
#   # the gene names are row names -- let's make them a column for easy display
#   tibble::rownames_to_column("Gene") %>%
#   # add a column for significance threshold results
#   dplyr::mutate(threshold = padj < 0.05) %>%
#   # sort by statistic -- the highest values will be genes with
#   # higher expression in RPL10 mutated samples
#   dplyr::arrange(dplyr::desc(log2FoldChange))

#   plotCounts(ddset, gene = "ENSMUSG00000026623", intgroup = "mutation_status")
# readr::write_tsv(
#   deseq_df,
#   file.path(
#     results_dir,
#     "SRP123625_diff_expr_results.tsv" # Replace with a relevant output file name
#   )
# )
# # We'll assign this as `volcano_plot`
# volcano_plot <- EnhancedVolcano::EnhancedVolcano(
#   deseq_df,
#   lab = deseq_df$Gene,
#   x = "log2FoldChange",
#   y = "padj",
#   pCutoff = 0.01 # Loosen the cutoff since we supplied corrected p-values
# )
# volcano_plot
# ggsave(
#   plot = volcano_plot,
#   file.path(plots_dir, "SRP123625_volcano_plot.png")
# ) # Replace with a plot name relevant to your data
# Load required packages
library(readr); library(dplyr); library(stringr); library(magrittr)
library(tibble); library(DESeq2); library(apeglm); library(EnhancedVolcano)
library(ggplot2)

# Paths (use project-relative paths)
metadata_path <- "data/SRP192714/metadata_SRP192714.tsv"
expr_path     <- "data/SRP192714/SRP192714.tsv"

# Read metadata and expression
metadata <- readr::read_tsv(metadata_path)

expression_df <- readr::read_tsv(expr_path) %>%
  tibble::column_to_rownames(var = "Gene")   # ensure 'Gene' is the column name

# Reorder expression columns to match the metadata samples (use all_of for safety)
# Make sure metadata$refinebio_accession_code contains the sample column names exactly.
expression_df <- expression_df %>% dplyr::select(all_of(metadata$refinebio_accession_code))

# Verify alignment
stopifnot(all.equal(colnames(expression_df), metadata$refinebio_accession_code))

# Create mutation_status from title (adjust regex to your data)
metadata <- metadata %>%
  dplyr::mutate(mutation_status = dplyr::case_when(
    stringr::str_detect(refinebio_title, "R98S") ~ "R98S",
    stringr::str_detect(refinebio_title, "WT") ~ "reference",
    TRUE ~ NA_character_
  ))

# Make sure mutation_status is a factor with desired levels
metadata$mutation_status <- factor(metadata$mutation_status, levels = c("reference", "R98S"))

# Filter low-count genes (sum across samples)
filtered_expression_df <- expression_df[rowSums(expression_df) >= 10, , drop = FALSE]

# Convert to integer matrix (DESeq2 expects counts)
gene_matrix <- round(as.matrix(filtered_expression_df))

# Ensure rownames are genes and column order matches metadata rows order
# (metadata rows must be in the same order as columns; if not, reorder metadata:)
metadata <- metadata[match(colnames(gene_matrix), metadata$refinebio_accession_code), ]

# Construct DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = gene_matrix,
                              colData = metadata,
                              design = ~ mutation_status)

dds <- DESeq(dds)

res <- results(dds)                 # raw results
res_shrunk <- lfcShrink(dds, coef = 2, type = "apeglm", res = res)  # requires apeglm

# Convert to tidy data.frame
deseq_df <- as.data.frame(res_shrunk) %>%
  tibble::rownames_to_column("Gene") %>%
  dplyr::mutate(threshold = ifelse(is.na(padj), FALSE, padj < 0.05)) %>%
  dplyr::arrange(dplyr::desc(log2FoldChange))

# Example count plot for a single gene
plotCounts(dds, gene = "ENSMUSG00000026623", intgroup = "mutation_status")  # adjust gene name

# Save results (ensure results_dir exists)
results_dir <- "results"
if (!dir.exists(results_dir)) dir.create(results_dir)
readr::write_tsv(deseq_df, file.path(results_dir, "SRP192714_diff_expr_results.tsv"))

# Volcano plot
volcano_plot <- EnhancedVolcano::EnhancedVolcano(
  deseq_df,
  lab = deseq_df$Gene,
  x = "log2FoldChange",
  y = "padj",
  pCutoff = 0.01
)

# Save volcano
plots_dir <- "plots"
if (!dir.exists(plots_dir)) dir.create(plots_dir)
ggsave(plot = volcano_plot, filename = file.path(plots_dir, "SRP192714_volcano_plot.png"), width = 8, height = 6)
```


# Section 3c
```{r}

```

# Section 3d
```{r}

```

# Section 3e
```{r}

```

# Section 4a
```{r}

```

# Section 4c
```{r}

```

# Section 5: William Le
```{r}

```

# Section 5: Nikhil Sangamkar
```{r}

```

# Section 5: Taylor Tillander
```{r}

```

# Section 5: Ibrahim Zbib
```{r}

```

# Section 6
```{r}

```

# Section 7
```{r}

```