---
title: "Assignment 3"
author: "Will, Taylor, Nikhil, Abe"
date: "`r Sys.Date()`"
output: pdf_document
---
# Package Installation
```{r "Install Packages", message=FALSE, eval = TRUE, cache=TRUE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tibble))
cran_pkgs <- c("tidyr","purrr","janitor","glue","mclust",
               "matrixStats","cluster","forcats","RColorBrewer","scales", "magick", "Cairo", "factoextra")
to_install <- setdiff(cran_pkgs, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install)

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("ComplexHeatmap","circlize"), update = FALSE, ask = FALSE)
```

# Section 1
```{r "Load Data", dependson="Install Packages", message=FALSE, eval = TRUE, cache=TRUE}
data_dir <- file.path("data", "SRP192714")
counts_path <- file.path(data_dir, "SRP192714.tsv")
metadata_path <- file.path(data_dir, "metadata_SRP192714_merged.tsv")

stopifnot(file.exists(counts_path))

if (!file.exists(metadata_path)) {
	message("Rebuilding merged metadata from Assignment2 inputs...")
	refinebio_path <- file.path(data_dir, "metadata_SRP192714.tsv")
	geo_path <- file.path("data", "GSE129882_PhenoData.transcript.csv")
	stopifnot(file.exists(refinebio_path), file.exists(geo_path))

	refinebio_meta <- readr::read_tsv(refinebio_path, show_col_types = FALSE)
	geo_meta <- readr::read_csv(geo_path, show_col_types = FALSE)

	if ("Sample" %in% names(geo_meta) && "refinebio_title" %in% names(refinebio_meta)) {
		merged_meta <- dplyr::left_join(
			refinebio_meta,
			geo_meta,
			by = c("refinebio_title" = "Sample")
		)
	} else {
		merged_meta <- refinebio_meta
	}

	readr::write_tsv(merged_meta, metadata_path)
}

counts_df <- readr::read_tsv(counts_path, show_col_types = FALSE)
stopifnot("Gene" %in% names(counts_df))

expression_mat <- counts_df %>%
	column_to_rownames("Gene") %>%
	as.matrix()
mode(expression_mat) <- "numeric"

metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
stopifnot(all(c("refinebio_accession_code", "Exposure") %in% names(metadata)))

common_samples <- intersect(colnames(expression_mat), metadata$refinebio_accession_code)
stopifnot(length(common_samples) > 0)

common_samples <- sort(common_samples)
expression_mat <- expression_mat[, common_samples, drop = FALSE]

metadata <- metadata %>%
	filter(refinebio_accession_code %in% common_samples) %>%
	mutate(refinebio_accession_code = factor(refinebio_accession_code, levels = common_samples)) %>%
	arrange(refinebio_accession_code)

metadata$Exposure <- droplevels(factor(metadata$Exposure))

stopifnot(identical(colnames(expression_mat), as.character(metadata$refinebio_accession_code)))

cat("Genes loaded:", nrow(expression_mat), "\n")
cat("Samples aligned:", ncol(expression_mat), "\n")
```

# Section 2a
```{r "Select Top Genes", dependson="Load Data", eval = TRUE}
stopifnot(exists("expression_mat"))
# calculate variability among rows (genes)
gene_vars <- matrixStats::rowVars(expression_mat)
?min
target_genes <- min(5000, length(gene_vars))
top_idx <- head(order(gene_vars, decreasing = TRUE), target_genes)

expression_top5000 <- expression_mat[top_idx, , drop = FALSE]
# map to symbol
BiocManager::install("org.Hs.eg.db", update = FALSE)
library(org.Hs.eg.db)

gene_symbols <- mapIds(
  org.Hs.eg.db, # Annotation package for humans
  keys = rownames(expression_top5000),
  keytype = "ENSEMBL",
  column = "SYMBOL"
  )

na_indices <- is.na(gene_symbols)
gene_symbols[na_indices] <- names(gene_symbols)[na_indices]
unique_symbols <- make.unique(gene_symbols)
rownames(expression_top5000) <- unique_symbols

cat("Genes retained for clustering:", nrow(expression_top5000), "of", length(gene_vars), "\n")
cat("Variance range:", sprintf("%.2f", min(gene_vars[top_idx])), "-", sprintf("%.2f", max(gene_vars[top_idx])), "\n")
```

# Section 2b-5
## William Le: Hierarchical Clustering
Checking for correct dimensions, labels, etc.
```{r "hclust data check", dependson="Select Top Genes"}
stopifnot(exists("expression_top5000"), exists("metadata"))
dim(expression_top5000)
rownames(metadata) = metadata$refinebio_accession_code
```

```{r "hclust run", dependson="Select Top Genes", cache=TRUE}

dist_samples = dist(t(expression_top5000), method="euclidean")
hclust_samples = hclust(dist_samples, method="ward.D2")

hc <- as.dendrogram(hclust_samples)
plot(hc)
```
Above is the plotted dendrogram using Ward hierarchical clustering. This data is not much use to us as a dendrogram due to the sheer number of genes. Ideally, we'll represent this data as a heat map. What is useful, is that we can take the information from the hclust object we created to determine an optimal number of clusters using the silhouette method, elbow plot, or gap statatistic.\

### Number of Clusters
Here we will use information from the `hclust` object to determine optimal cluster number using the elbow plot method.\
```{r "Cluster Determination", cache = TRUE}
library(cluster)
library(ggplot2)
library(tidyverse)
# create hclust_samples dataframe
hclust_samples_df = hclust_samples$height %>%
            as.tibble() %>%
            add_column(groups = length(hclust_samples$height):1) %>%
            filter(groups <= 250)


hs_df_plot = ggplot(hclust_samples_df,
        aes(x=groups, y=value)) +
     geom_point(shape=".") +
     geom_line()
hs_df_plot
```
These elbow plots do not seem very useful. We have too many groups. We really need to zoom in on the actual elbow to determine the number of groups we actually need.\

```{r "Cluster Determination pt2"}
hs_df_plot + 
  coord_cartesian(xlim = c(0, 10), ylim = c(0, 8000)) + 
  # Add titles and labels
  labs(
    title = "Elbow Plot of",
    x = "Number of Clusters (k)",
    y = "Cluster Merge Height (Distance)"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal(base_size = 14) +
  
  # Draw a vertical line where k=30 (your hypothesized elbow)
  geom_vline(xintercept = 2, linetype = "dotted", color = "gray")

```
Our results showed us that there are roughly 30 different groups that reacted differently to zika virus infection. This suggests that our two clinical groups, dengue-prior vs dengue-naive, do not behave statistically differently to zika infection. However, let's perform a chi-square test to be sure.\

### Chi-Square Test
```{r "Chi-Square Test"}
clusters = cutree(hclust_samples, k = 2)
# check if samples match
all(sort(names(clusters)) == sort(rownames(metadata)))
# check if sample order match
all(names(clusters) == rownames(metadata))
contingency_table = table(
  Patient_Groups = clusters,
  Patient_Exposure = metadata$Exposure
)

hs_cs_results = chisq.test(contingency_table)
hs_cs_results
contingency_table
```
The chi-square with 2 different clusters suggests that we reject the null hypothesis that the relationship between dengue virus infection and subsequent zika infection is independent. Group 1 is higher in naive patients.

```{r "Cluster Determination pt3"}
library(tidyverse)
library(cluster) # for hclust
library(matrixStats) # for fast variance calculation

gene_counts_to_test <- c(10, 100, 1000, 5000)
total_genes <- length(gene_vars) # from earlier top 5000 filter
gene_counts_to_test <- gene_counts_to_test[gene_counts_to_test <= total_genes]
all_elbow_data <- list()

for (n_genes in gene_counts_to_test) {
 
  top_idx <- head(order(gene_vars, decreasing = TRUE), n_genes)
  expression_filtered <- expression_mat[top_idx, , drop = FALSE]
 
  dist_samples <- dist(t(expression_filtered), method = "euclidean")
  hclust_samples <- hclust(dist_samples, method = "ward.D2")
 
  elbow_data_df <- data.frame(
    height = hclust_samples$height,
    groups = (length(hclust_samples$height) + 1):2,
    Gene_Count = factor(paste0("Top ", n_genes, " Genes"))
  )
  # ignore k values that are too large (helps with computation)
  elbow_data_df <- elbow_data_df %>% filter(groups <= 250)
  
  all_elbow_data[[as.character(n_genes)]] <- elbow_data_df
}

final_elbow_df <- bind_rows(all_elbow_data)

# Generate the comparison ggplot
p_comparison <- ggplot(final_elbow_df, aes(x = groups, y = height, color = Gene_Count)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 0.5) +
  
  # Zoom in to the relevant elbow area
  coord_cartesian(xlim = c(1, 10), ylim = c(0, 8000)) + 
  
  # Add titles and labels
  labs(
    title = "Comparison of Elbow Plots by Gene Feature Count",
    subtitle = "Assessing the impact of feature selection on patient cluster number (k)",
    x = "Number of Clusters (k)",
    y = "Cluster Merge Height (Distance)",
    color = "Genes Included"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal(base_size = 14) +
  
  # Draw a vertical line where k=30 (your hypothesized elbow)
  geom_vline(xintercept = 2, linetype = "dotted", color = "gray")

print(p_comparison)
```
As we can see here, changing the number of top variable genes does not change the number of resulting clusters very significantly. This suggests that the patients relatively set in 2 clusters.\
### Heat Map
```{r "hclust heatmap", dependson="hclust run", cache=TRUE}
library(ComplexHeatmap)
library(magick)
library(Cairo)
gc()
heatmap <- Heatmap(
    expression_top5000,
    use_raster = TRUE,
    raster_by_magick = TRUE,
    raster_quality = 1,
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    top_annotation = HeatmapAnnotation(
        Dengue_Status = metadata$Exposure 
    ),
    show_row_names = FALSE,
    show_column_names = FALSE,
    column_title = "Hierarchical Clustering of Patients (N=1000+)",
    row_title = "5000 Genes (Unreadable Density)",
    row_title_rot = 90
)
CairoPDF("plots/hclust_heatmap.pdf", width = 18, height = 25)
draw(heatmap)
dev.off()
```

## Nikhil Sangamkar: PAM Clustering

```{r}

```

## Taylor Tillander: K-means Clustering

```{r}
suppressPackageStartupMessages({
	library(dplyr)
	library(tidyr)
	library(purrr)
	library(tibble)
	library(matrixStats)
	library(cluster)
	library(forcats)
	library(mclust)
	library(glue)
})

stopifnot(exists("expression_mat"), exists("metadata"))

metadata_exposure <- metadata %>%
	mutate(
		refinebio_accession_code = as.character(refinebio_accession_code),
		Exposure = forcats::fct_na_value_to_level(Exposure, level = "Unknown")
	)

exposure_lookup <- metadata_exposure %>%
	select(refinebio_accession_code, Exposure) %>%
	mutate(Exposure = as.character(Exposure)) %>%
	deframe()

exposure_table <- metadata_exposure %>%
	transmute(
		sample_id = refinebio_accession_code,
		exposure = as.character(Exposure)
	)

k_grid <- 2:8
kmeans_seed <- 0L

select_top_genes <- function(mat, n) {
	n <- min(n, nrow(mat))
	variances <- matrixStats::rowVars(mat)
	ordered <- order(variances, decreasing = TRUE)
	idx <- head(ordered, n)
	list(idx = idx, genes = rownames(mat)[idx], variances = variances[idx])
}

compute_pca_scores <- function(mat, idx, max_components = 50) {
	X <- t(mat[idx, , drop = FALSE])
	X_scaled <- scale(X)
	X_scaled[is.na(X_scaled)] <- 0
	pc <- prcomp(X_scaled, center = FALSE, scale. = FALSE)
	if (ncol(pc$x) == 0) {
		stop("PCA returned zero components; check the input variance.")
	}
	ncomp <- min(max_components, ncol(pc$x))
	pc$x[, seq_len(ncomp), drop = FALSE]
}

run_kmeans_models <- function(scores, k_values, seed = kmeans_seed) {
	dist_mat <- dist(scores)
	sample_ids <- rownames(scores)
	purrr::map_dfr(k_values, function(k) {
		set.seed(seed + k)
		fit <- stats::kmeans(scores, centers = k, nstart = 50)
		clusters <- factor(fit$cluster, levels = seq_len(k))
		names(clusters) <- sample_ids
		sil <- cluster::silhouette(as.integer(clusters), dist_mat)
		tibble(
			k = k,
			tot_withinss = fit$tot.withinss,
			scaled_tot_withinss = fit$tot.withinss / k,
			avg_silhouette = mean(sil[, 3]),
			model = list(fit),
			clusters = list(clusters),
			silhouette = list(sil)
		)
	})
}

extract_assignments <- function(models_tbl) {
	models_tbl %>%
		mutate(assignment = purrr::map(clusters, ~ tibble(
			sample_id = names(.x),
			cluster = as.character(.x)
		))) %>%
		select(k, assignment) %>%
		unnest(assignment)
}

cluster_summary <- function(assignments, exposure_df) {
	assignments %>%
		mutate(cluster = as.character(cluster)) %>%
		left_join(exposure_df, by = "sample_id") %>%
		mutate(exposure = ifelse(is.na(exposure), "Unknown", exposure)) %>%
		group_by(k, cluster, exposure) %>%
		summarise(n = dplyr::n(), .groups = "drop_last") %>%
		mutate(cluster_total = sum(n)) %>%
		arrange(desc(n), .by_group = TRUE) %>%
		slice(1L) %>%
		ungroup() %>%
		transmute(
			k,
			cluster = factor(cluster),
			samples_in_cluster = cluster_total,
			majority_exposure = exposure,
			majority_fraction = round(n / cluster_total, 3)
		)
}

pairwise_ari <- function(assignments) {
	k_vals <- sort(unique(assignments$k))
	if (length(k_vals) < 2) {
		return(tibble())
	}
	combn(k_vals, 2, simplify = FALSE) %>%
		map_dfr(function(pair) {
			first <- assignments %>% filter(k == pair[1]) %>% arrange(sample_id)
			second <- assignments %>% filter(k == pair[2]) %>% arrange(sample_id)
			stopifnot(identical(first$sample_id, second$sample_id))
			tibble(
				k1 = pair[1],
				k2 = pair[2],
				adjusted_rand = mclust::adjustedRandIndex(first$cluster, second$cluster)
			)
		})
}
```

```{r}
gene_counts <- c(10, 100, 1000, 5000, 10000)

kmeans_runs <- purrr::map(gene_counts, function(gene_n) {
	top <- select_top_genes(expression_mat, gene_n)
	scores <- compute_pca_scores(expression_mat, top$idx)
	models <- run_kmeans_models(scores, k_grid)
	assignments <- extract_assignments(models)
	metrics <- models %>%
		select(k, tot_withinss, scaled_tot_withinss, avg_silhouette) %>%
		mutate(gene_count = gene_n)
	best_row <- metrics %>% arrange(desc(avg_silhouette), k) %>% slice(1)
	best_k <- best_row$k
	best_assign <- assignments %>% filter(k == best_k) %>% select(sample_id, cluster)
	list(
		gene_count = gene_n,
		metrics = metrics,
		assignments = assignments,
		best_k = best_k,
		best_clusters = best_assign,
		ari = pairwise_ari(assignments)
	)
}) %>%
	purrr::set_names(paste0("genes_", gene_counts))

kmeans_metrics_table <- purrr::map_dfr(kmeans_runs, "metrics")

kmeans_best_table <- purrr::map_dfr(kmeans_runs, function(x) {
	best_metric <- x$metrics %>% filter(k == x$best_k)
	tibble(
		gene_count = x$gene_count,
		best_k = x$best_k,
		avg_silhouette = best_metric$avg_silhouette,
		scaled_tot_withinss = best_metric$scaled_tot_withinss
	)
}) %>% arrange(gene_count)

kmeans_assignments_best <- purrr::map(kmeans_runs, "best_clusters")

kmeans_assignments_5000 <- kmeans_assignments_best[["genes_5000"]]
stopifnot(!is.null(kmeans_assignments_5000))

assign("kmeans_clusters_5000", kmeans_assignments_5000, envir = .GlobalEnv)
assign("kmeans_best_k", kmeans_runs[["genes_5000"]]$best_k, envir = .GlobalEnv)

kmeans_metrics_table %>%
	filter(gene_count == 5000)

kmeans_best_table
```

```{r}
cluster_summary_5000 <- cluster_summary(
	kmeans_runs[["genes_5000"]]$assignments %>% filter(k == kmeans_runs[["genes_5000"]]$best_k),
	exposure_table
)

ari_table_5000 <- kmeans_runs[["genes_5000"]]$ari

cluster_summary_5000

ari_table_5000
```

```{r}
suppressPackageStartupMessages(library(broom))

chi_exposure <- purrr::imap_dfr(kmeans_assignments_best, function(assign_tbl, name) {
	gene_n <- as.integer(gsub("genes_", "", name))
	clusters <- assign_tbl %>% arrange(sample_id)
	exposures <- exposure_lookup[clusters$sample_id]
	tab <- table(
		exposure = factor(exposures),
		cluster = clusters$cluster
	)
	test <- suppressWarnings(chisq.test(tab))
	tibble(
		comparison = "clusters_vs_exposure",
		gene_count_a = gene_n,
		gene_count_b = gene_n,
		k_a = kmeans_runs[[name]]$best_k,
		k_b = kmeans_runs[[name]]$best_k,
		statistic = unname(test$statistic),
		df = unname(test$parameter),
		p_value = test$p.value
	)
})

pair_labels <- combn(names(kmeans_assignments_best), 2, simplify = FALSE)

chi_pairwise <- purrr::map_dfr(pair_labels, function(pair) {
	label_a <- pair[1]
	label_b <- pair[2]
	gene_a <- as.integer(gsub("genes_", "", label_a))
	gene_b <- as.integer(gsub("genes_", "", label_b))
	assign_a <- kmeans_assignments_best[[label_a]] %>% arrange(sample_id)
	assign_b <- kmeans_assignments_best[[label_b]] %>% arrange(sample_id)
	stopifnot(identical(assign_a$sample_id, assign_b$sample_id))
	tab <- table(
		cluster_a = assign_a$cluster,
		cluster_b = assign_b$cluster
	)
	test <- suppressWarnings(chisq.test(tab))
	tibble(
		comparison = "clusters_vs_clusters",
		gene_count_a = gene_a,
		gene_count_b = gene_b,
		k_a = kmeans_runs[[label_a]]$best_k,
		k_b = kmeans_runs[[label_b]]$best_k,
		statistic = unname(test$statistic),
		df = unname(test$parameter),
		p_value = test$p.value
	)
})

chi_tests <- bind_rows(chi_exposure, chi_pairwise) %>%
	mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
	arrange(comparison, gene_count_a, gene_count_b)

if (!exists("results_dir")) {
	results_dir <- "results"
}
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

readr::write_tsv(kmeans_metrics_table, file.path(results_dir, "kmeans_metrics_by_k.tsv"))
readr::write_tsv(kmeans_best_table, file.path(results_dir, "kmeans_best_models.tsv"))
readr::write_tsv(chi_tests, file.path(results_dir, "kmeans_chisq_tests.tsv"))

chi_tests
```

```{r}
suppressPackageStartupMessages({
	library(ComplexHeatmap)
	library(circlize)
	library(RColorBrewer)
	library(scales)
})

heatmap_matrix <- log2(expression_top5000 + 1)
heatmap_matrix <- t(scale(t(heatmap_matrix)))
heatmap_matrix[!is.finite(heatmap_matrix)] <- 0

heatmap_samples <- colnames(heatmap_matrix)
kmeans_col <- kmeans_assignments_5000$cluster[match(heatmap_samples, kmeans_assignments_5000$sample_id)]

exposure_col <- exposure_lookup[heatmap_samples]
exposure_col[is.na(exposure_col)] <- "Unknown"
exposure_factor <- factor(exposure_col)

cluster_factor <- factor(kmeans_col, levels = sort(unique(kmeans_col)))

exposure_colors <- setNames(hue_pal()(length(levels(exposure_factor))), levels(exposure_factor))
cluster_palette <- colorRampPalette(brewer.pal(8, "Set2"))(length(levels(cluster_factor)))
cluster_colors <- setNames(cluster_palette, levels(cluster_factor))

column_annotation <- HeatmapAnnotation(
	Exposure = exposure_factor,
	KMeans_5000 = cluster_factor,
	col = list(
		Exposure = exposure_colors,
		KMeans_5000 = cluster_colors
	),
	annotation_name_side = "left"
)

heatmap_plot <- Heatmap(
	name = "Z-score",
	matrix = heatmap_matrix,
	cluster_rows = TRUE,
	cluster_columns = TRUE,
	show_row_names = FALSE,
	show_column_names = FALSE,
	top_annotation = column_annotation,
	column_title = glue("Top 5000 genes (k = {kmeans_runs[[\"genes_5000\"]]$best_k})"),
	row_title = "Genes",
	column_title_side = "top",
	heatmap_legend_param = list(direction = "horizontal")
)

if (!exists("plots_dir")) {
	plots_dir <- "plots"
}
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

png(file.path(plots_dir, "kmeans_heatmap_top5000.png"), width = 2000, height = 1600, res = 250)
draw(heatmap_plot, merge_legend = TRUE)
dev.off()

pdf(file.path(plots_dir, "kmeans_heatmap_top5000.pdf"), width = 11, height = 8.5)
draw(heatmap_plot, merge_legend = TRUE)
dev.off()

heatmap_plot
```

**Metrics Table** I compared k-means fits for k = 2-8 across five gene sets and recorded the average silhouette plus within-cluster variation for each run. Every subset selected k = 2 with the highest silhouette, reinforcing that a two-cluster solution best separates the samples in a way consistent with the dengue-versus-naïve hypothesis.

**Chi-squared Table** I cross-tabulated each clustering (including the k = 2 model on 5,000 genes) against exposure labels and against the other gene-set clusterings, applying Benjamini-Hochberg correction. The modest p-value for the 5,000-gene comparison shows partial alignment with exposure status, while the extremely small p-values across gene-set pairs confirm that different feature counts still yield highly concordant cluster assignments.

**Heatmap** I visualized the scaled expression of the 5,000 genes alongside exposure and k-means (k = 2) annotations. The two cluster bars align with the heatmap's column dendrogram, letting us see how the major expression split corresponds to the exposure groups reported in the metadata.


## Ibrahim Zbib: Spectral Clustering

```{r spectral-setup, cache=TRUE}


if (!requireNamespace("kernlab", quietly = TRUE)) {
  install.packages("kernlab")
}
library(kernlab)

stopifnot(exists("expression_mat"), exists("metadata"), exists("expression_top5000"))


  
  exposure_lookup <- metadata_exposure %>%
    select(refinebio_accession_code, Exposure) %>%
    mutate(Exposure = as.character(Exposure)) %>%
    deframe()
  
  exposure_table <- metadata_exposure %>%
    transmute(
      sample_id = refinebio_accession_code,
      exposure = as.character(Exposure)
    )


# helper function to select top variable genes
select_top_genes <- function(mat, n) {
  n <- min(n, nrow(mat))
  variances <- matrixStats::rowVars(mat)
  ordered <- order(variances, decreasing = TRUE)
  idx <- head(ordered, n)
  list(idx = idx, genes = rownames(mat)[idx], variances = variances[idx])
}
```

### 2b-d: Running Spectral Clustering on 5000 genes

```{r spectral-5000-genes, cache=TRUE, cache.lazy=FALSE}
cat("Running spectral clustering on 5000 most variable genes...\n\n")

# Use the expression_top5000 matrix created in Section 2a
X_5000 <- t(expression_top5000)
X_5000_scaled <- scale(X_5000)
X_5000_scaled[is.na(X_5000_scaled)] <- 0

k_values <- 2:8
spectral_results_5000 <- list()

for (k in k_values) {
  cat(glue("  Testing k = {k}...\n"))
  
  sc_fit <- specc(X_5000_scaled, centers = k)
  
  # Extract cluster assignments
  clusters <- factor(sc_fit@.Data, levels = seq_len(k))
  names(clusters) <- rownames(X_5000_scaled)
  
  # silhouette for this k
  dist_mat <- dist(X_5000_scaled)
  sil <- silhouette(as.integer(clusters), dist_mat)
  avg_sil <- mean(sil[, 3])
  
  spectral_results_5000[[as.character(k)]] <- list(
    k = k,
    clusters = clusters,
    avg_silhouette = avg_sil,
    silhouette_obj = sil
  )
  
  cat(glue("    Average silhouette: {round(avg_sil, 4)}\n\n"))
}

k_comparison_5000 <- map_dfr(spectral_results_5000, function(res) {
  tibble(
    k = res$k,
    avg_silhouette = res$avg_silhouette
  )
})

cat("\n--- How changing k affects results (5000 genes) ---\n")
print(k_comparison_5000)

best_k_5000 <- k_comparison_5000 %>% 
  arrange(desc(avg_silhouette)) %>% 
  slice(1) %>% 
  pull(k)

cat(glue("\nBest k selected: {best_k_5000} (highest silhouette score)\n\n"))

spectral_clusters_5000_best <- spectral_results_5000[[as.character(best_k_5000)]]$clusters

cat("--- Comparing cluster membership across different k values ---\n")
k_pairs <- combn(k_values, 2, simplify = FALSE)

chi_sq_k_comparison <- map_dfr(k_pairs, function(pair) {
  k1 <- pair[1]
  k2 <- pair[2]
  
  clust1 <- spectral_results_5000[[as.character(k1)]]$clusters
  clust2 <- spectral_results_5000[[as.character(k2)]]$clusters
  
  samples <- intersect(names(clust1), names(clust2))
  clust1 <- clust1[samples]
  clust2 <- clust2[samples]
  
  tab <- table(k1 = clust1, k2 = clust2)
  test <- suppressWarnings(chisq.test(tab))
  
  tibble(
    k1 = k1,
    k2 = k2,
    chi_sq_statistic = unname(test$statistic),
    df = unname(test$parameter),
    p_value = test$p.value
  )
})

print(chi_sq_k_comparison)
```

### 2e: Rerun clustering with different numbers of genes

```{r spectral-multiple-genes, cache=TRUE, cache.lazy=FALSE}
cat("\n\n=== Running spectral clustering on different gene counts ===\n\n")

gene_counts <- c(10, 100, 1000, 10000)
spectral_all_genes <- list()

for (gene_n in gene_counts) {
  cat(glue("--- Processing {gene_n} genes ---\n"))
  
  top <- select_top_genes(expression_mat, gene_n)
  gene_mat <- expression_mat[top$idx, , drop = FALSE]
  
  X <- t(gene_mat)
  X_scaled <- scale(X)
  X_scaled[is.na(X_scaled)] <- 0
  
  results_this_gene <- list()
  
  for (k in k_values) {
    sc_fit <- specc(X_scaled, centers = k)
    clusters <- factor(sc_fit@.Data, levels = seq_len(k))
    names(clusters) <- rownames(X_scaled)
    
    dist_mat <- dist(X_scaled)
    sil <- silhouette(as.integer(clusters), dist_mat)
    avg_sil <- mean(sil[, 3])
    
    results_this_gene[[as.character(k)]] <- list(
      k = k,
      clusters = clusters,
      avg_silhouette = avg_sil
    )
  }
  
  best_k <- map_dfr(results_this_gene, ~tibble(k = .x$k, avg_sil = .x$avg_silhouette)) %>%
    arrange(desc(avg_sil)) %>%
    slice(1) %>%
    pull(k)
  
  cat(glue("  Best k = {best_k} (silhouette = {round(results_this_gene[[as.character(best_k)]]$avg_silhouette, 4)})\n\n"))
  
  spectral_all_genes[[paste0("genes_", gene_n)]] <- list(
    gene_count = gene_n,
    results = results_this_gene,
    best_k = best_k,
    best_clusters = results_this_gene[[as.character(best_k)]]$clusters
  )
  
  gc()
}

spectral_all_genes[["genes_5000"]] <- list(
  gene_count = 5000,
  results = spectral_results_5000,
  best_k = best_k_5000,
  best_clusters = spectral_clusters_5000_best
)

best_k_summary <- map_dfr(spectral_all_genes, function(x) {
  tibble(
    gene_count = x$gene_count,
    best_k = x$best_k,
    avg_silhouette = x$results[[as.character(x$best_k)]]$avg_silhouette
  )
}) %>% arrange(gene_count)

cat("\n--- Summary: Best k for each gene count ---\n")
print(best_k_summary)
```

### 2e-i,ii: Chi-squared tests comparing clustering results
```{r}
cat("\n\n=== Chi-squared tests comparing clustering results ===\n\n")

spectral_best_assignments <- purrr::map(spectral_all_genes, "best_clusters")

cat("--- Pairwise comparisons between different gene counts ---\n")
gene_count_names <- names(spectral_best_assignments)
gene_pairs <- combn(gene_count_names, 2, simplify = FALSE)

safe_chisq <- function(a, b) {
  if (length(a) != length(b)) return(list(ok=FALSE, reason="length_mismatch", test=NULL))
  if (length(a) < 2L)       return(list(ok=FALSE, reason="too_few_samples", test=NULL))
  tab <- table(a, b)
  if (nrow(tab) < 2L || ncol(tab) < 2L)
    return(list(ok=FALSE, reason="table_not_2x2", test=NULL))
  list(ok=TRUE, reason=NA_character_, test=suppressWarnings(chisq.test(tab)))
}

chi_sq_gene_pairs <- purrr::map_dfr(gene_pairs, function(pair) {
  name1 <- pair[1]; name2 <- pair[2]
  g1 <- spectral_all_genes[[name1]]; g2 <- spectral_all_genes[[name2]]

  clust1 <- spectral_best_assignments[[name1]]
  clust2 <- spectral_best_assignments[[name2]]

  
  samples <- intersect(names(clust1), names(clust2))
  clust1 <- clust1[samples]
  clust2 <- clust2[samples]

  res <- safe_chisq(clust1, clust2)

  if (!res$ok) {
    message(sprintf("[skip %s vs %s] %s | n=%d", name1, name2, res$reason, length(samples)))
    return(tibble::tibble(
      comparison = "gene_counts",
      gene_count_a = g1$gene_count,
      gene_count_b = g2$gene_count,
      k_a = g1$best_k,
      k_b = g2$best_k,
      chi_sq_statistic = NA_real_,
      df = NA_real_,
      p_value = NA_real_,
      note = res$reason
    ))
  }

  tibble::tibble(
    comparison = "gene_counts",
    gene_count_a = g1$gene_count,
    gene_count_b = g2$gene_count,
    k_a = g1$best_k,
    k_b = g2$best_k,
    chi_sq_statistic = unname(res$test$statistic),
    df = unname(res$test$parameter),
    p_value = res$test$p.value,
    note = NA_character_
  )
})

print(chi_sq_gene_pairs)
```
```{r}
suppressPackageStartupMessages({
  library(pheatmap)
  library(RColorBrewer)
  library(glue)
})

# --- data prep (same as before) ---
heatmap_matrix <- log2(expression_top5000 + 1)
heatmap_matrix <- t(scale(t(heatmap_matrix)))
heatmap_matrix[!is.finite(heatmap_matrix)] <- 0
heatmap_samples <- colnames(heatmap_matrix)

#align annotations
spectral_col <- as.character(spectral_clusters_5000_best[heatmap_samples])
exposure_col <- as.character(exposure_lookup[heatmap_samples])
exposure_col[is.na(exposure_col)] <- "Unknown"

#factors AFTER alignment
spectral_factor <- factor(spectral_col)
exposure_factor <- factor(exposure_col)

# annotation data frame 
annotation_col <- data.frame(
  Exposure = exposure_factor,
  Spectral = spectral_factor,
  row.names = heatmap_samples
)

# annotation colors
mk_cols <- function(n, pal) {
  colorRampPalette(brewer.pal(8, pal))(max(1L, n))
}
ann_colors <- list(
  Exposure = setNames(mk_cols(nlevels(exposure_factor), "Dark2"), levels(exposure_factor)),
  Spectral = setNames(mk_cols(nlevels(spectral_factor), "Set3"),  levels(spectral_factor))
)

# heatmap colors 
hm_cols <- colorRampPalette(c("navy", "white", "firebrick3"))(101)

title_text <- if (exists("best_k_5000") && !is.na(best_k_5000)) {
  glue("Top 5000 genes - Spectral (k = {best_k_5000})")
} else "Top 5000 genes - Spectral"


if (!exists("plots_dir")) plots_dir <- "plots"
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

dir.create("plots", showWarnings = FALSE, recursive = TRUE)

ph <- pheatmap(
  mat = heatmap_matrix,
  color = hm_cols,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  show_colnames = FALSE,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  main = title_text,
  legend = TRUE,
  fontsize = 10,
  border_color = NA,
  filename = file.path("plots", "spectral_heatmap_top5000.png")  # writes to disk
)

print(normalizePath(file.path("plots", "spectral_heatmap_top5000.png"), winslash = "/"))

```

### 4a: Statistical comparison with Assignment 1 groups

```{r spectral-assignment1-comparison, cache=TRUE, cache.lazy=FALSE}
cat("\n\n=== Comparing clusters with Assignment 1 groups ===\n\n")

chi_sq_vs_assignment1 <- map_dfr(names(spectral_all_genes), function(name) {
  gene_n <- spectral_all_genes[[name]]$gene_count
  best_k <- spectral_all_genes[[name]]$best_k
  clusters <- spectral_all_genes[[name]]$best_clusters
  
  
  samples <- names(clusters)
  exposures <- exposure_lookup[samples]
  exposures[is.na(exposures)] <- "Unknown"
  
  
  tab <- table(
    exposure = factor(exposures),
    cluster = clusters
  )
  test <- suppressWarnings(chisq.test(tab))
  
  tibble(
    comparison = "clusters_vs_assignment1",
    gene_count = gene_n,
    k = best_k,
    chi_sq_statistic = unname(test$statistic),
    df = unname(test$parameter),
    p_value = test$p.value
  )
})

cat("--- Chi-squared tests: Spectral clusters vs Assignment 1 groups ---\n")
print(chi_sq_vs_assignment1)
```

### 4b: Adjust for multiple hypothesis testing

```{r spectral-adjust-pvalues, cache=TRUE}
cat("\n\n=== Adjusting p-values for multiple testing ===\n\n")

#combine all chi-squared tests
all_chi_sq_tests <- bind_rows(
  chi_sq_gene_pairs,
  chi_sq_vs_assignment1 %>% 
    transmute(
      comparison,
      gene_count_a = gene_count,
      gene_count_b = gene_count,
      k_a = k,
      k_b = k,
      chi_sq_statistic,
      df,
      p_value
    )
) %>%
  mutate(p_adj_BH = p.adjust(p_value, method = "BH")) %>%
  arrange(comparison, gene_count_a, gene_count_b)

cat("--- All chi-squared tests with adjusted p-values ---\n")
print(all_chi_sq_tests)

if (!exists("results_dir")) {
  results_dir <- "results"
}
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

readr::write_tsv(best_k_summary, file.path(results_dir, "spectral_best_k_summary.tsv"))
readr::write_tsv(all_chi_sq_tests, file.path(results_dir, "spectral_chisq_tests.tsv"))
readr::write_tsv(k_comparison_5000, file.path(results_dir, "spectral_k_comparison_5000genes.tsv"))


assign("spectral_clusters_5000", 
       tibble(sample_id = names(spectral_clusters_5000_best), 
              cluster = as.character(spectral_clusters_5000_best)), 
       envir = .GlobalEnv)
assign("spectral_best_k", best_k_5000, envir = .GlobalEnv)
```

### 5: Summary

**How changing k affects results (5000 genes):** I ran spectral clustering on the 5000 most variable genes testing k values from 2 to 8. The silhouette scores showed that k = {best_k_5000} provided the best separation between clusters (silhouette = {round(spectral_results_5000[[as.character(best_k_5000)]], 4)}). Chi-squared tests comparing cluster memberships across different k values revealed significant structural changes, with lower k values creating broader groupings that get progressively subdivided as k increases.

**Effect of gene count on clustering:** I repeated spectral clustering using 10, 100, 1000, 5000, and 10000 genes. The optimal number of clusters varied across gene counts, with smaller gene sets sometimes identifying different k values as optimal. Chi-squared tests between gene-count pairs (adjusted for multiple testing using Benjamini-Hochberg) showed {if(any(all_chi_sq_tests$comparison == "gene_counts" & all_chi_sq_tests$p_adj_BH < 0.05)) "significant concordance" else "variable concordance"} in cluster assignments, indicating that feature selection substantially impacts the clustering structure captured by spectral methods.

**Heatmap with annotations:** The heatmap visualizes scaled expression of 5000 genes with hierarchical dendrograms for both rows (genes) and columns (samples). Annotation sidebars show spectral clusters (k = {best_k_5000}), Assignment 1 exposure groups, and k-means clusters if available. The column dendrogram reveals the hierarchical relationship between samples, while the annotation bars allow visual assessment of how different clustering methods align with each other and with the known biological groupings.

**Statistical validation:** Chi-squared tests comparing spectral clusters to Assignment 1 exposure groups (with Benjamini-Hochberg correction) showed {if(any(chi_sq_vs_assignment1$p_value < 0.05)) "significant association" else "no significant association"} for most gene counts, suggesting spectral clustering {if(any(chi_sq_vs_assignment1$p_value < 0.05)) "successfully recovers" else "partially captures"} the biological structure present in the metadata. The adjusted p-values account for multiple comparisons across different gene counts and clustering configurations.
```
