# Assignment 3

# Package Installation
```{r}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tibble))
cran_pkgs <- c("tidyr","purrr","janitor","glue","mclust",
               "matrixStats","cluster","forcats","RColorBrewer","scales")
to_install <- setdiff(cran_pkgs, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install)

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("ComplexHeatmap","circlize"), update = FALSE, ask = FALSE)
```

# Section 1
```{r}
data_dir <- file.path("data", "SRP192714")
counts_path <- file.path(data_dir, "SRP192714.tsv")
metadata_path <- file.path(data_dir, "metadata_SRP192714_merged.tsv")

stopifnot(file.exists(counts_path))

if (!file.exists(metadata_path)) {
	message("Rebuilding merged metadata from Assignment2 inputs...")
	refinebio_path <- file.path(data_dir, "metadata_SRP192714.tsv")
	geo_path <- file.path("data", "GSE129882_PhenoData.transcript.csv")
	stopifnot(file.exists(refinebio_path), file.exists(geo_path))

	refinebio_meta <- readr::read_tsv(refinebio_path, show_col_types = FALSE)
	geo_meta <- readr::read_csv(geo_path, show_col_types = FALSE)

	if ("Sample" %in% names(geo_meta) && "refinebio_title" %in% names(refinebio_meta)) {
		merged_meta <- dplyr::left_join(
			refinebio_meta,
			geo_meta,
			by = c("refinebio_title" = "Sample")
		)
	} else {
		merged_meta <- refinebio_meta
	}

	readr::write_tsv(merged_meta, metadata_path)
}

counts_df <- readr::read_tsv(counts_path, show_col_types = FALSE)
stopifnot("Gene" %in% names(counts_df))

expression_mat <- counts_df %>%
	column_to_rownames("Gene") %>%
	as.matrix()
mode(expression_mat) <- "numeric"

metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
stopifnot(all(c("refinebio_accession_code", "Exposure") %in% names(metadata)))

common_samples <- intersect(colnames(expression_mat), metadata$refinebio_accession_code)
stopifnot(length(common_samples) > 0)

common_samples <- sort(common_samples)
expression_mat <- expression_mat[, common_samples, drop = FALSE]

metadata <- metadata %>%
	filter(refinebio_accession_code %in% common_samples) %>%
	mutate(refinebio_accession_code = factor(refinebio_accession_code, levels = common_samples)) %>%
	arrange(refinebio_accession_code)

metadata$Exposure <- droplevels(factor(metadata$Exposure))

stopifnot(identical(colnames(expression_mat), as.character(metadata$refinebio_accession_code)))

cat("Genes loaded:", nrow(expression_mat), "\n")
cat("Samples aligned:", ncol(expression_mat), "\n")
```

# Section 2a
```{r}
stopifnot(exists("expression_mat"))

gene_vars <- matrixStats::rowVars(expression_mat)

target_genes <- min(5000, length(gene_vars))
top_idx <- head(order(gene_vars, decreasing = TRUE), target_genes)

expression_top5000 <- expression_mat[top_idx, , drop = FALSE]

cat("Genes retained for clustering:", nrow(expression_top5000), "of", length(gene_vars), "\n")
cat("Variance range:", sprintf("%.2f", min(gene_vars[top_idx])), "-", sprintf("%.2f", max(gene_vars[top_idx])), "\n")
```

# Section 2b-5
## William Le: Hierarchical Clustering
```{r}

```

## Nikhil Sangamkar: PAM Clustering

```{r}

```

## Taylor Tillander: K-means Clustering

```{r}
suppressPackageStartupMessages({
	library(dplyr)
	library(tidyr)
	library(purrr)
	library(tibble)
	library(matrixStats)
	library(cluster)
	library(forcats)
	library(mclust)
	library(glue)
})

stopifnot(exists("expression_mat"), exists("metadata"))

metadata_exposure <- metadata %>%
	mutate(
		refinebio_accession_code = as.character(refinebio_accession_code),
		Exposure = forcats::fct_na_value_to_level(Exposure, level = "Unknown")
	)

exposure_lookup <- metadata_exposure %>%
	select(refinebio_accession_code, Exposure) %>%
	mutate(Exposure = as.character(Exposure)) %>%
	deframe()

exposure_table <- metadata_exposure %>%
	transmute(
		sample_id = refinebio_accession_code,
		exposure = as.character(Exposure)
	)

k_grid <- 2:8
kmeans_seed <- 0L

select_top_genes <- function(mat, n) {
	n <- min(n, nrow(mat))
	variances <- matrixStats::rowVars(mat)
	ordered <- order(variances, decreasing = TRUE)
	idx <- head(ordered, n)
	list(idx = idx, genes = rownames(mat)[idx], variances = variances[idx])
}

compute_pca_scores <- function(mat, idx, max_components = 50) {
	X <- t(mat[idx, , drop = FALSE])
	X_scaled <- scale(X)
	X_scaled[is.na(X_scaled)] <- 0
	pc <- prcomp(X_scaled, center = FALSE, scale. = FALSE)
	if (ncol(pc$x) == 0) {
		stop("PCA returned zero components; check the input variance.")
	}
	ncomp <- min(max_components, ncol(pc$x))
	pc$x[, seq_len(ncomp), drop = FALSE]
}

run_kmeans_models <- function(scores, k_values, seed = kmeans_seed) {
	dist_mat <- dist(scores)
	sample_ids <- rownames(scores)
	purrr::map_dfr(k_values, function(k) {
		set.seed(seed + k)
		fit <- stats::kmeans(scores, centers = k, nstart = 50)
		clusters <- factor(fit$cluster, levels = seq_len(k))
		names(clusters) <- sample_ids
		sil <- cluster::silhouette(as.integer(clusters), dist_mat)
		tibble(
			k = k,
			tot_withinss = fit$tot.withinss,
			scaled_tot_withinss = fit$tot.withinss / k,
			avg_silhouette = mean(sil[, 3]),
			model = list(fit),
			clusters = list(clusters),
			silhouette = list(sil)
		)
	})
}

extract_assignments <- function(models_tbl) {
	models_tbl %>%
		mutate(assignment = purrr::map(clusters, ~ tibble(
			sample_id = names(.x),
			cluster = as.character(.x)
		))) %>%
		select(k, assignment) %>%
		unnest(assignment)
}

cluster_summary <- function(assignments, exposure_df) {
	assignments %>%
		mutate(cluster = as.character(cluster)) %>%
		left_join(exposure_df, by = "sample_id") %>%
		mutate(exposure = ifelse(is.na(exposure), "Unknown", exposure)) %>%
		group_by(k, cluster, exposure) %>%
		summarise(n = dplyr::n(), .groups = "drop_last") %>%
		mutate(cluster_total = sum(n)) %>%
		arrange(desc(n), .by_group = TRUE) %>%
		slice(1L) %>%
		ungroup() %>%
		transmute(
			k,
			cluster = factor(cluster),
			samples_in_cluster = cluster_total,
			majority_exposure = exposure,
			majority_fraction = round(n / cluster_total, 3)
		)
}

pairwise_ari <- function(assignments) {
	k_vals <- sort(unique(assignments$k))
	if (length(k_vals) < 2) {
		return(tibble())
	}
	combn(k_vals, 2, simplify = FALSE) %>%
		map_dfr(function(pair) {
			first <- assignments %>% filter(k == pair[1]) %>% arrange(sample_id)
			second <- assignments %>% filter(k == pair[2]) %>% arrange(sample_id)
			stopifnot(identical(first$sample_id, second$sample_id))
			tibble(
				k1 = pair[1],
				k2 = pair[2],
				adjusted_rand = mclust::adjustedRandIndex(first$cluster, second$cluster)
			)
		})
}
```

```{r}
gene_counts <- c(10, 100, 1000, 5000, 10000)

kmeans_runs <- purrr::map(gene_counts, function(gene_n) {
	top <- select_top_genes(expression_mat, gene_n)
	scores <- compute_pca_scores(expression_mat, top$idx)
	models <- run_kmeans_models(scores, k_grid)
	assignments <- extract_assignments(models)
	metrics <- models %>%
		select(k, tot_withinss, scaled_tot_withinss, avg_silhouette) %>%
		mutate(gene_count = gene_n)
	best_row <- metrics %>% arrange(desc(avg_silhouette), k) %>% slice(1)
	best_k <- best_row$k
	best_assign <- assignments %>% filter(k == best_k) %>% select(sample_id, cluster)
	list(
		gene_count = gene_n,
		metrics = metrics,
		assignments = assignments,
		best_k = best_k,
		best_clusters = best_assign,
		ari = pairwise_ari(assignments)
	)
}) %>%
	purrr::set_names(paste0("genes_", gene_counts))

kmeans_metrics_table <- purrr::map_dfr(kmeans_runs, "metrics")

kmeans_best_table <- purrr::map_dfr(kmeans_runs, function(x) {
	best_metric <- x$metrics %>% filter(k == x$best_k)
	tibble(
		gene_count = x$gene_count,
		best_k = x$best_k,
		avg_silhouette = best_metric$avg_silhouette,
		scaled_tot_withinss = best_metric$scaled_tot_withinss
	)
}) %>% arrange(gene_count)

kmeans_assignments_best <- purrr::map(kmeans_runs, "best_clusters")

kmeans_assignments_5000 <- kmeans_assignments_best[["genes_5000"]]
stopifnot(!is.null(kmeans_assignments_5000))

assign("kmeans_clusters_5000", kmeans_assignments_5000, envir = .GlobalEnv)
assign("kmeans_best_k", kmeans_runs[["genes_5000"]]$best_k, envir = .GlobalEnv)

kmeans_metrics_table %>%
	filter(gene_count == 5000)

kmeans_best_table
```

```{r}
cluster_summary_5000 <- cluster_summary(
	kmeans_runs[["genes_5000"]]$assignments %>% filter(k == kmeans_runs[["genes_5000"]]$best_k),
	exposure_table
)

ari_table_5000 <- kmeans_runs[["genes_5000"]]$ari

cluster_summary_5000

ari_table_5000
```

```{r}
suppressPackageStartupMessages(library(broom))

chi_exposure <- purrr::imap_dfr(kmeans_assignments_best, function(assign_tbl, name) {
	gene_n <- as.integer(gsub("genes_", "", name))
	clusters <- assign_tbl %>% arrange(sample_id)
	exposures <- exposure_lookup[clusters$sample_id]
	tab <- table(
		exposure = factor(exposures),
		cluster = clusters$cluster
	)
	test <- suppressWarnings(chisq.test(tab))
	tibble(
		comparison = "clusters_vs_exposure",
		gene_count_a = gene_n,
		gene_count_b = gene_n,
		k_a = kmeans_runs[[name]]$best_k,
		k_b = kmeans_runs[[name]]$best_k,
		statistic = unname(test$statistic),
		df = unname(test$parameter),
		p_value = test$p.value
	)
})

pair_labels <- combn(names(kmeans_assignments_best), 2, simplify = FALSE)

chi_pairwise <- purrr::map_dfr(pair_labels, function(pair) {
	label_a <- pair[1]
	label_b <- pair[2]
	gene_a <- as.integer(gsub("genes_", "", label_a))
	gene_b <- as.integer(gsub("genes_", "", label_b))
	assign_a <- kmeans_assignments_best[[label_a]] %>% arrange(sample_id)
	assign_b <- kmeans_assignments_best[[label_b]] %>% arrange(sample_id)
	stopifnot(identical(assign_a$sample_id, assign_b$sample_id))
	tab <- table(
		cluster_a = assign_a$cluster,
		cluster_b = assign_b$cluster
	)
	test <- suppressWarnings(chisq.test(tab))
	tibble(
		comparison = "clusters_vs_clusters",
		gene_count_a = gene_a,
		gene_count_b = gene_b,
		k_a = kmeans_runs[[label_a]]$best_k,
		k_b = kmeans_runs[[label_b]]$best_k,
		statistic = unname(test$statistic),
		df = unname(test$parameter),
		p_value = test$p.value
	)
})

chi_tests <- bind_rows(chi_exposure, chi_pairwise) %>%
	mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
	arrange(comparison, gene_count_a, gene_count_b)

if (!exists("results_dir")) {
	results_dir <- "results"
}
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

readr::write_tsv(kmeans_metrics_table, file.path(results_dir, "kmeans_metrics_by_k.tsv"))
readr::write_tsv(kmeans_best_table, file.path(results_dir, "kmeans_best_models.tsv"))
readr::write_tsv(chi_tests, file.path(results_dir, "kmeans_chisq_tests.tsv"))

chi_tests
```

```{r}
suppressPackageStartupMessages({
	library(ComplexHeatmap)
	library(circlize)
	library(RColorBrewer)
	library(scales)
})

heatmap_matrix <- log2(expression_top5000 + 1)
heatmap_matrix <- t(scale(t(heatmap_matrix)))
heatmap_matrix[!is.finite(heatmap_matrix)] <- 0

heatmap_samples <- colnames(heatmap_matrix)
kmeans_col <- kmeans_assignments_5000$cluster[match(heatmap_samples, kmeans_assignments_5000$sample_id)]

exposure_col <- exposure_lookup[heatmap_samples]
exposure_col[is.na(exposure_col)] <- "Unknown"
exposure_factor <- factor(exposure_col)

cluster_factor <- factor(kmeans_col, levels = sort(unique(kmeans_col)))

exposure_colors <- setNames(hue_pal()(length(levels(exposure_factor))), levels(exposure_factor))
cluster_palette <- colorRampPalette(brewer.pal(8, "Set2"))(length(levels(cluster_factor)))
cluster_colors <- setNames(cluster_palette, levels(cluster_factor))

column_annotation <- HeatmapAnnotation(
	Exposure = exposure_factor,
	KMeans_5000 = cluster_factor,
	col = list(
		Exposure = exposure_colors,
		KMeans_5000 = cluster_colors
	),
	annotation_name_side = "left"
)

heatmap_plot <- Heatmap(
	name = "Z-score",
	matrix = heatmap_matrix,
	cluster_rows = TRUE,
	cluster_columns = TRUE,
	show_row_names = FALSE,
	show_column_names = FALSE,
	top_annotation = column_annotation,
	column_title = glue("Top 5000 genes (k = {kmeans_runs[[\"genes_5000\"]]$best_k})"),
	row_title = "Genes",
	column_title_side = "top",
	heatmap_legend_param = list(direction = "horizontal")
)

if (!exists("plots_dir")) {
	plots_dir <- "plots"
}
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

png(file.path(plots_dir, "kmeans_heatmap_top5000.png"), width = 2000, height = 1600, res = 250)
draw(heatmap_plot, merge_legend = TRUE)
dev.off()

pdf(file.path(plots_dir, "kmeans_heatmap_top5000.pdf"), width = 11, height = 8.5)
draw(heatmap_plot, merge_legend = TRUE)
dev.off()

heatmap_plot
```

**Metrics Table** I compared k-means fits for k = 2-8 across five gene sets and recorded the average silhouette plus within-cluster variation for each run. Every subset selected k = 2 with the highest silhouette, reinforcing that a two-cluster solution best separates the samples in a way consistent with the dengue-versus-naïve hypothesis.

**Chi-squared Table** I cross-tabulated each clustering (including the k = 2 model on 5,000 genes) against exposure labels and against the other gene-set clusterings, applying Benjamini-Hochberg correction. The modest p-value for the 5,000-gene comparison shows partial alignment with exposure status, while the extremely small p-values across gene-set pairs confirm that different feature counts still yield highly concordant cluster assignments.

**Heatmap** I visualized the scaled expression of the 5,000 genes alongside exposure and k-means (k = 2) annotations. The two cluster bars align with the heatmap's column dendrogram, letting us see how the major expression split corresponds to the exposure groups reported in the metadata.


## Ibrahim Zbib: Gaussian Clustering

```{r}

```