---
title: "Assignment 3"
author: "Will, Taylor, Nikhil, Abe"
date: "`r Sys.Date()`"
output: pdf_document
---
# Package Installation
```{r "Install Packages", message=FALSE, eval = TRUE, cache=TRUE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tibble))
cran_pkgs <- c("tidyr","purrr","janitor","glue","mclust",
               "matrixStats","cluster","forcats","RColorBrewer","scales", "magick", "Cairo", "factoextra")
to_install <- setdiff(cran_pkgs, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install)

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("ComplexHeatmap","circlize"), update = FALSE, ask = FALSE)
```

# Section 1
```{r "Load Data", dependson="Install Packages", message=FALSE, eval = TRUE, cache=TRUE}
data_dir <- file.path("data", "SRP192714")
counts_path <- file.path(data_dir, "SRP192714.tsv")
metadata_path <- file.path(data_dir, "metadata_SRP192714_merged.tsv")

stopifnot(file.exists(counts_path))

if (!file.exists(metadata_path)) {
	message("Rebuilding merged metadata from Assignment2 inputs...")
	refinebio_path <- file.path(data_dir, "metadata_SRP192714.tsv")
	geo_path <- file.path("data", "GSE129882_PhenoData.transcript.csv")
	stopifnot(file.exists(refinebio_path), file.exists(geo_path))

	refinebio_meta <- readr::read_tsv(refinebio_path, show_col_types = FALSE)
	geo_meta <- readr::read_csv(geo_path, show_col_types = FALSE)

	if ("Sample" %in% names(geo_meta) && "refinebio_title" %in% names(refinebio_meta)) {
		merged_meta <- dplyr::left_join(
			refinebio_meta,
			geo_meta,
			by = c("refinebio_title" = "Sample")
		)
	} else {
		merged_meta <- refinebio_meta
	}

	readr::write_tsv(merged_meta, metadata_path)
}

counts_df <- readr::read_tsv(counts_path, show_col_types = FALSE)
stopifnot("Gene" %in% names(counts_df))

expression_mat <- counts_df %>%
	column_to_rownames("Gene") %>%
	as.matrix()
mode(expression_mat) <- "numeric"

metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
stopifnot(all(c("refinebio_accession_code", "Exposure") %in% names(metadata)))

common_samples <- intersect(colnames(expression_mat), metadata$refinebio_accession_code)
stopifnot(length(common_samples) > 0)

common_samples <- sort(common_samples)
expression_mat <- expression_mat[, common_samples, drop = FALSE]

metadata <- metadata %>%
	filter(refinebio_accession_code %in% common_samples) %>%
	mutate(refinebio_accession_code = factor(refinebio_accession_code, levels = common_samples)) %>%
	arrange(refinebio_accession_code)

metadata$Exposure <- droplevels(factor(metadata$Exposure))

stopifnot(identical(colnames(expression_mat), as.character(metadata$refinebio_accession_code)))

cat("Genes loaded:", nrow(expression_mat), "\n")
cat("Samples aligned:", ncol(expression_mat), "\n")
```

# Section 2a
```{r "Select Top Genes", dependson="Load Data", eval = TRUE}
stopifnot(exists("expression_mat"))
# calculate variability among rows (genes)
gene_vars <- matrixStats::rowVars(expression_mat)
?min
target_genes <- min(5000, length(gene_vars))
top_idx <- head(order(gene_vars, decreasing = TRUE), target_genes)

expression_top5000 <- expression_mat[top_idx, , drop = FALSE]
# map to symbol
BiocManager::install("org.Hs.eg.db", update = FALSE)
library(org.Hs.eg.db)

gene_symbols <- mapIds(
  org.Hs.eg.db, # Annotation package for humans
  keys = rownames(expression_top5000),
  keytype = "ENSEMBL",
  column = "SYMBOL"
  )

na_indices <- is.na(gene_symbols)
gene_symbols[na_indices] <- names(gene_symbols)[na_indices]
unique_symbols <- make.unique(gene_symbols)
rownames(expression_top5000) <- unique_symbols

cat("Genes retained for clustering:", nrow(expression_top5000), "of", length(gene_vars), "\n")
cat("Variance range:", sprintf("%.2f", min(gene_vars[top_idx])), "-", sprintf("%.2f", max(gene_vars[top_idx])), "\n")
```

# Section 2b-5
## William Le: Hierarchical Clustering
Checking for correct dimensions, labels, etc.
```{r "hclust data check", dependson="Select Top Genes"}
stopifnot(exists("expression_top5000"), exists("metadata"))
dim(expression_top5000)
rownames(metadata) = metadata$refinebio_accession_code
```

```{r "hclust run", dependson="Select Top Genes", cache=TRUE}

dist_samples = dist(t(expression_top5000), method="euclidean")
hclust_samples = hclust(dist_samples, method="ward.D2")

hc <- as.dendrogram(hclust_samples)
plot(hc)
```
Above is the plotted dendrogram using Ward hierarchical clustering. This data is not much use to us as a dendrogram due to the sheer number of genes. Ideally, we'll represent this data as a heat map. What is useful, is that we can take the information from the hclust object we created to determine an optimal number of clusters using the silhouette method, elbow plot, or gap statatistic.\

### Number of Clusters
Here we will use information from the `hclust` object to determine optimal cluster number using the elbow plot method.\
```{r "Cluster Determination", cache = TRUE}
library(cluster)
library(ggplot2)
library(tidyverse)
# create hclust_samples dataframe
hclust_samples_df = hclust_samples$height %>%
            as.tibble() %>%
            add_column(groups = length(hclust_samples$height):1)


hs_df_plot = ggplot(hclust_samples_df,
        aes(x=groups, y=value)) +
     geom_point(shape=".") +
     geom_line()
hs_df_plot
```
These elbow plots do not seem very useful. We have too many groups. We really need to zoom in on the actual elbow to determine the number of groups we actually need.\
```{r "Cluster Determination pt2"}
hs_df_plot + 
  coord_cartesian(xlim = c(0, 250), ylim = c(0, 2000)) + 
  geom_abline(intercept = 1760, slope = -20, color = "red")
```
Our results showed us that there are around 35 different groups that reacted differently to zika virus infection. This suggests that our two clinical groups, dengue-prior vs dengue-naive, do not behave statistically differently to zika infection. However, let's perform a chi-square test to be sure.\

### Chi-Square Test
```{r "Chi-Square Test"}
clusters = cutree(hclust_samples, k = 35)
# check if samples match
all(sort(names(clusters)) == sort(rownames(metadata)))
# check if sample order match
all(names(clusters) == rownames(metadata))
contingency_table = table(
  Patient_Groups = clusters,
  Patient_Exposure = metadata$Exposure
)

hs_cs_results = chisq.test(contingency_table)
hs_cs_results
```


### Heat Map
```{r "hclust heatmap", dependson="hclust run", cache=TRUE}
library(ComplexHeatmap)
library(magick)
library(Cairo)
gc()
heatmap <- Heatmap(
    expression_top5000,
    use_raster = TRUE,
    raster_by_magick = TRUE,
    raster_quality = 1,
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    top_annotation = HeatmapAnnotation(
        Dengue_Status = metadata$Exposure 
    ),
    show_row_names = FALSE,
    show_column_names = FALSE,
    column_title = "Hierarchical Clustering of Patients (N=1000+)",
    row_title = "5000 Genes (Unreadable Density)",
    row_title_rot = 90
)
CairoPDF("plots/hclust_heatmap.pdf", width = 18, height = 25)
draw(heatmap)
dev.off()
```

## Nikhil Sangamkar: PAM Clustering

```{r}

```

## Taylor Tillander: K-means Clustering

```{r}
suppressPackageStartupMessages({
	library(dplyr)
	library(tidyr)
	library(purrr)
	library(tibble)
	library(matrixStats)
	library(cluster)
	library(forcats)
	library(mclust)
	library(glue)
})

stopifnot(exists("expression_mat"), exists("metadata"))

metadata_exposure <- metadata %>%
	mutate(
		refinebio_accession_code = as.character(refinebio_accession_code),
		Exposure = forcats::fct_na_value_to_level(Exposure, level = "Unknown")
	)

exposure_lookup <- metadata_exposure %>%
	select(refinebio_accession_code, Exposure) %>%
	mutate(Exposure = as.character(Exposure)) %>%
	deframe()

exposure_table <- metadata_exposure %>%
	transmute(
		sample_id = refinebio_accession_code,
		exposure = as.character(Exposure)
	)

k_grid <- 2:8
kmeans_seed <- 0L

select_top_genes <- function(mat, n) {
	n <- min(n, nrow(mat))
	variances <- matrixStats::rowVars(mat)
	ordered <- order(variances, decreasing = TRUE)
	idx <- head(ordered, n)
	list(idx = idx, genes = rownames(mat)[idx], variances = variances[idx])
}

compute_pca_scores <- function(mat, idx, max_components = 50) {
	X <- t(mat[idx, , drop = FALSE])
	X_scaled <- scale(X)
	X_scaled[is.na(X_scaled)] <- 0
	pc <- prcomp(X_scaled, center = FALSE, scale. = FALSE)
	if (ncol(pc$x) == 0) {
		stop("PCA returned zero components; check the input variance.")
	}
	ncomp <- min(max_components, ncol(pc$x))
	pc$x[, seq_len(ncomp), drop = FALSE]
}

run_kmeans_models <- function(scores, k_values, seed = kmeans_seed) {
	dist_mat <- dist(scores)
	sample_ids <- rownames(scores)
	purrr::map_dfr(k_values, function(k) {
		set.seed(seed + k)
		fit <- stats::kmeans(scores, centers = k, nstart = 50)
		clusters <- factor(fit$cluster, levels = seq_len(k))
		names(clusters) <- sample_ids
		sil <- cluster::silhouette(as.integer(clusters), dist_mat)
		tibble(
			k = k,
			tot_withinss = fit$tot.withinss,
			scaled_tot_withinss = fit$tot.withinss / k,
			avg_silhouette = mean(sil[, 3]),
			model = list(fit),
			clusters = list(clusters),
			silhouette = list(sil)
		)
	})
}

extract_assignments <- function(models_tbl) {
	models_tbl %>%
		mutate(assignment = purrr::map(clusters, ~ tibble(
			sample_id = names(.x),
			cluster = as.character(.x)
		))) %>%
		select(k, assignment) %>%
		unnest(assignment)
}

cluster_summary <- function(assignments, exposure_df) {
	assignments %>%
		mutate(cluster = as.character(cluster)) %>%
		left_join(exposure_df, by = "sample_id") %>%
		mutate(exposure = ifelse(is.na(exposure), "Unknown", exposure)) %>%
		group_by(k, cluster, exposure) %>%
		summarise(n = dplyr::n(), .groups = "drop_last") %>%
		mutate(cluster_total = sum(n)) %>%
		arrange(desc(n), .by_group = TRUE) %>%
		slice(1L) %>%
		ungroup() %>%
		transmute(
			k,
			cluster = factor(cluster),
			samples_in_cluster = cluster_total,
			majority_exposure = exposure,
			majority_fraction = round(n / cluster_total, 3)
		)
}

pairwise_ari <- function(assignments) {
	k_vals <- sort(unique(assignments$k))
	if (length(k_vals) < 2) {
		return(tibble())
	}
	combn(k_vals, 2, simplify = FALSE) %>%
		map_dfr(function(pair) {
			first <- assignments %>% filter(k == pair[1]) %>% arrange(sample_id)
			second <- assignments %>% filter(k == pair[2]) %>% arrange(sample_id)
			stopifnot(identical(first$sample_id, second$sample_id))
			tibble(
				k1 = pair[1],
				k2 = pair[2],
				adjusted_rand = mclust::adjustedRandIndex(first$cluster, second$cluster)
			)
		})
}
```

```{r}
gene_counts <- c(10, 100, 1000, 5000, 10000)

kmeans_runs <- purrr::map(gene_counts, function(gene_n) {
	top <- select_top_genes(expression_mat, gene_n)
	scores <- compute_pca_scores(expression_mat, top$idx)
	models <- run_kmeans_models(scores, k_grid)
	assignments <- extract_assignments(models)
	metrics <- models %>%
		select(k, tot_withinss, scaled_tot_withinss, avg_silhouette) %>%
		mutate(gene_count = gene_n)
	best_row <- metrics %>% arrange(desc(avg_silhouette), k) %>% slice(1)
	best_k <- best_row$k
	best_assign <- assignments %>% filter(k == best_k) %>% select(sample_id, cluster)
	list(
		gene_count = gene_n,
		metrics = metrics,
		assignments = assignments,
		best_k = best_k,
		best_clusters = best_assign,
		ari = pairwise_ari(assignments)
	)
}) %>%
	purrr::set_names(paste0("genes_", gene_counts))

kmeans_metrics_table <- purrr::map_dfr(kmeans_runs, "metrics")

kmeans_best_table <- purrr::map_dfr(kmeans_runs, function(x) {
	best_metric <- x$metrics %>% filter(k == x$best_k)
	tibble(
		gene_count = x$gene_count,
		best_k = x$best_k,
		avg_silhouette = best_metric$avg_silhouette,
		scaled_tot_withinss = best_metric$scaled_tot_withinss
	)
}) %>% arrange(gene_count)

kmeans_assignments_best <- purrr::map(kmeans_runs, "best_clusters")

kmeans_assignments_5000 <- kmeans_assignments_best[["genes_5000"]]
stopifnot(!is.null(kmeans_assignments_5000))

assign("kmeans_clusters_5000", kmeans_assignments_5000, envir = .GlobalEnv)
assign("kmeans_best_k", kmeans_runs[["genes_5000"]]$best_k, envir = .GlobalEnv)

kmeans_metrics_table %>%
	filter(gene_count == 5000)

kmeans_best_table
```

```{r}
cluster_summary_5000 <- cluster_summary(
	kmeans_runs[["genes_5000"]]$assignments %>% filter(k == kmeans_runs[["genes_5000"]]$best_k),
	exposure_table
)

ari_table_5000 <- kmeans_runs[["genes_5000"]]$ari

cluster_summary_5000

ari_table_5000
```

```{r}
suppressPackageStartupMessages(library(broom))

chi_exposure <- purrr::imap_dfr(kmeans_assignments_best, function(assign_tbl, name) {
	gene_n <- as.integer(gsub("genes_", "", name))
	clusters <- assign_tbl %>% arrange(sample_id)
	exposures <- exposure_lookup[clusters$sample_id]
	tab <- table(
		exposure = factor(exposures),
		cluster = clusters$cluster
	)
	test <- suppressWarnings(chisq.test(tab))
	tibble(
		comparison = "clusters_vs_exposure",
		gene_count_a = gene_n,
		gene_count_b = gene_n,
		k_a = kmeans_runs[[name]]$best_k,
		k_b = kmeans_runs[[name]]$best_k,
		statistic = unname(test$statistic),
		df = unname(test$parameter),
		p_value = test$p.value
	)
})

pair_labels <- combn(names(kmeans_assignments_best), 2, simplify = FALSE)

chi_pairwise <- purrr::map_dfr(pair_labels, function(pair) {
	label_a <- pair[1]
	label_b <- pair[2]
	gene_a <- as.integer(gsub("genes_", "", label_a))
	gene_b <- as.integer(gsub("genes_", "", label_b))
	assign_a <- kmeans_assignments_best[[label_a]] %>% arrange(sample_id)
	assign_b <- kmeans_assignments_best[[label_b]] %>% arrange(sample_id)
	stopifnot(identical(assign_a$sample_id, assign_b$sample_id))
	tab <- table(
		cluster_a = assign_a$cluster,
		cluster_b = assign_b$cluster
	)
	test <- suppressWarnings(chisq.test(tab))
	tibble(
		comparison = "clusters_vs_clusters",
		gene_count_a = gene_a,
		gene_count_b = gene_b,
		k_a = kmeans_runs[[label_a]]$best_k,
		k_b = kmeans_runs[[label_b]]$best_k,
		statistic = unname(test$statistic),
		df = unname(test$parameter),
		p_value = test$p.value
	)
})

chi_tests <- bind_rows(chi_exposure, chi_pairwise) %>%
	mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
	arrange(comparison, gene_count_a, gene_count_b)

if (!exists("results_dir")) {
	results_dir <- "results"
}
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

readr::write_tsv(kmeans_metrics_table, file.path(results_dir, "kmeans_metrics_by_k.tsv"))
readr::write_tsv(kmeans_best_table, file.path(results_dir, "kmeans_best_models.tsv"))
readr::write_tsv(chi_tests, file.path(results_dir, "kmeans_chisq_tests.tsv"))

chi_tests
```

```{r}
suppressPackageStartupMessages({
	library(ComplexHeatmap)
	library(circlize)
	library(RColorBrewer)
	library(scales)
})

heatmap_matrix <- log2(expression_top5000 + 1)
heatmap_matrix <- t(scale(t(heatmap_matrix)))
heatmap_matrix[!is.finite(heatmap_matrix)] <- 0

heatmap_samples <- colnames(heatmap_matrix)
kmeans_col <- kmeans_assignments_5000$cluster[match(heatmap_samples, kmeans_assignments_5000$sample_id)]

exposure_col <- exposure_lookup[heatmap_samples]
exposure_col[is.na(exposure_col)] <- "Unknown"
exposure_factor <- factor(exposure_col)

cluster_factor <- factor(kmeans_col, levels = sort(unique(kmeans_col)))

exposure_colors <- setNames(hue_pal()(length(levels(exposure_factor))), levels(exposure_factor))
cluster_palette <- colorRampPalette(brewer.pal(8, "Set2"))(length(levels(cluster_factor)))
cluster_colors <- setNames(cluster_palette, levels(cluster_factor))

column_annotation <- HeatmapAnnotation(
	Exposure = exposure_factor,
	KMeans_5000 = cluster_factor,
	col = list(
		Exposure = exposure_colors,
		KMeans_5000 = cluster_colors
	),
	annotation_name_side = "left"
)

heatmap_plot <- Heatmap(
	name = "Z-score",
	matrix = heatmap_matrix,
	cluster_rows = TRUE,
	cluster_columns = TRUE,
	show_row_names = FALSE,
	show_column_names = FALSE,
	top_annotation = column_annotation,
	column_title = glue("Top 5000 genes (k = {kmeans_runs[[\"genes_5000\"]]$best_k})"),
	row_title = "Genes",
	column_title_side = "top",
	heatmap_legend_param = list(direction = "horizontal")
)

if (!exists("plots_dir")) {
	plots_dir <- "plots"
}
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

png(file.path(plots_dir, "kmeans_heatmap_top5000.png"), width = 2000, height = 1600, res = 250)
draw(heatmap_plot, merge_legend = TRUE)
dev.off()

pdf(file.path(plots_dir, "kmeans_heatmap_top5000.pdf"), width = 11, height = 8.5)
draw(heatmap_plot, merge_legend = TRUE)
dev.off()

heatmap_plot
```

**Metrics Table** I compared k-means fits for k = 2-8 across five gene sets and recorded the average silhouette plus within-cluster variation for each run. Every subset selected k = 2 with the highest silhouette, reinforcing that a two-cluster solution best separates the samples in a way consistent with the dengue-versus-naïve hypothesis.

**Chi-squared Table** I cross-tabulated each clustering (including the k = 2 model on 5,000 genes) against exposure labels and against the other gene-set clusterings, applying Benjamini-Hochberg correction. The modest p-value for the 5,000-gene comparison shows partial alignment with exposure status, while the extremely small p-values across gene-set pairs confirm that different feature counts still yield highly concordant cluster assignments.

**Heatmap** I visualized the scaled expression of the 5,000 genes alongside exposure and k-means (k = 2) annotations. The two cluster bars align with the heatmap's column dendrogram, letting us see how the major expression split corresponds to the exposure groups reported in the metadata.


## Ibrahim Zbib: Spectral Clustering

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(grid)
  library(glue)
})
suppressPackageStartupMessages({
  library(purrr)
  library(dplyr)
  library(tibble)
})

plots_dir <- "plots"
results_dir <- "results"
dir.create(plots_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

save_table_png <- function(df, path, base_height = 400, row_px = 28, min_width = 1200){
  h <- max(base_height, nrow(df) * row_px + 120)
  w <- min_width
  g <- gridExtra::tableGrob(df)
  png(path, width = w, height = h, res = 150)
  grid::grid.newpage()
  grid::grid.draw(g)
  dev.off()
}

save_gg_png <- function(plot, path, width = 10, height = 7, dpi = 300){
  ggplot2::ggsave(filename = path, plot = plot, width = width, height = height, dpi = dpi, limitsize = FALSE)
}
```


```{r spectral-setup, cache=TRUE}


if (!requireNamespace("kernlab", quietly = TRUE)) {
  install.packages("kernlab")
}
library(kernlab)

stopifnot(exists("expression_mat"), exists("metadata"), exists("expression_top5000"))


  
  #exposure_lookup <- metadata_exposure %>%
   # select(refinebio_accession_code, Exposure) %>%
    #mutate(Exposure = as.character(Exposure)) %>%
    #deframe()
  
#  exposure_table <- metadata_exposure %>%
 #   transmute(
  #    sample_id = refinebio_accession_code,
   #   exposure = as.character(Exposure)
    #)


# helper function to select top variable genes
select_top_genes <- function(mat, n) {
  n <- min(n, nrow(mat))
  variances <- matrixStats::rowVars(mat)
  ordered <- order(variances, decreasing = TRUE)
  idx <- head(ordered, n)
  list(idx = idx, genes = rownames(mat)[idx], variances = variances[idx])
}
```

### 2b-d: Running Spectral Clustering on 5000 genes

```{r spectral-5000-genes, cache=TRUE, cache.lazy=FALSE}
cat("Running spectral clustering on 5000 most variable genes...\n\n")

# Use the expression_top5000 matrix created in Section 2a
X_5000 <- t(expression_top5000)
X_5000_scaled <- scale(X_5000)
X_5000_scaled[is.na(X_5000_scaled)] <- 0

k_values <- 2:8
spectral_results_5000 <- list()

for (k in k_values) {
  cat(glue("  Testing k = {k}...\n"))
  
  sc_fit <- specc(X_5000_scaled, centers = k)
  
  # Extract cluster assignments
  clusters <- factor(sc_fit@.Data, levels = seq_len(k))
  names(clusters) <- rownames(X_5000_scaled)
  
  # silhouette for this k
  dist_mat <- dist(X_5000_scaled)
  sil <- silhouette(as.integer(clusters), dist_mat)
  avg_sil <- mean(sil[, 3])
  
  spectral_results_5000[[as.character(k)]] <- list(
    k = k,
    clusters = clusters,
    avg_silhouette = avg_sil,
    silhouette_obj = sil
  )
  
  cat(glue("    Average silhouette: {round(avg_sil, 4)}\n\n"))
}

k_comparison_5000 <- map_dfr(spectral_results_5000, function(res) {
  tibble(
    k = res$k,
    avg_silhouette = res$avg_silhouette
  )
})

```
```{r}
cat("\n--- How changing k affects results (5000 genes) ---\n")
print(k_comparison_5000)

# robust: base indexing; no dplyr::slice
best_k_5000 <- k_comparison_5000$k[which.max(k_comparison_5000$avg_silhouette)]

cat(glue("\nBest k selected: {best_k_5000} (highest silhouette score)\n\n"))

spectral_clusters_5000_best <- spectral_results_5000[[as.character(best_k_5000)]]$clusters

cat("--- Comparing cluster membership across different k values ---\n")
k_pairs <- combn(k_values, 2, simplify = FALSE)

chi_sq_k_comparison <- map_dfr(k_pairs, function(pair) {
  k1 <- pair[1]
  k2 <- pair[2]
  
  clust1 <- spectral_results_5000[[as.character(k1)]]$clusters
  clust2 <- spectral_results_5000[[as.character(k2)]]$clusters
  
  samples <- intersect(names(clust1), names(clust2))
  clust1 <- clust1[samples]
  clust2 <- clust2[samples]
  
  tab <- table(k1 = clust1, k2 = clust2)
  test <- suppressWarnings(chisq.test(tab))
  
  tibble(
    k1 = k1,
    k2 = k2,
    chi_sq_statistic = unname(test$statistic),
    df = unname(test$parameter),
    p_value = test$p.value
  )
})
```

### 2e: Rerun clustering with different numbers of genes

```{r spectral-multiple-genes, cache=TRUE, cache.lazy=FALSE}
cat("\n\n=== Running spectral clustering on different gene counts ===\n\n")

gene_counts <- c(10, 100, 1000, 10000)
spectral_all_genes <- list()

for (gene_n in gene_counts) {
  cat(glue("--- Processing {gene_n} genes ---\n"))
  
  top <- select_top_genes(expression_mat, gene_n)
  gene_mat <- expression_mat[top$idx, , drop = FALSE]
  
  X <- t(gene_mat)
  X_scaled <- scale(X)
  X_scaled[is.na(X_scaled)] <- 0
  
  results_this_gene <- list()
  
  for (k in k_values) {
    sc_fit <- specc(X_scaled, centers = k)
    clusters <- factor(sc_fit@.Data, levels = seq_len(k))
    names(clusters) <- rownames(X_scaled)
    
    dist_mat <- dist(X_scaled)
    sil <- silhouette(as.integer(clusters), dist_mat)
    avg_sil <- mean(sil[, 3])
    
    results_this_gene[[as.character(k)]] <- list(
      k = k,
      clusters = clusters,
      avg_silhouette = avg_sil
    )
  }
  
 best_df <- map_dfr(results_this_gene, ~tibble(k = .x$k, avg_sil = .x$avg_silhouette))
best_k  <- best_df$k[which.max(best_df$avg_sil)]
  
  cat(glue("  Best k = {best_k} (silhouette = {round(results_this_gene[[as.character(best_k)]]$avg_silhouette, 4)})\n\n"))
  
  spectral_all_genes[[paste0("genes_", gene_n)]] <- list(
    gene_count = gene_n,
    results = results_this_gene,
    best_k = best_k,
    best_clusters = results_this_gene[[as.character(best_k)]]$clusters
  )
  
  gc()
}

spectral_all_genes[["genes_5000"]] <- list(
  gene_count = 5000,
  results = spectral_results_5000,
  best_k = best_k_5000,
  best_clusters = spectral_clusters_5000_best
)

best_k_summary <- map_dfr(spectral_all_genes, function(x) {
  tibble(
    gene_count = x$gene_count,
    best_k = x$best_k,
    avg_silhouette = x$results[[as.character(x$best_k)]]$avg_silhouette
  )
}) %>% arrange(gene_count)

cat("\n--- Summary: Best k for each gene count ---\n")
print(best_k_summary)
```

### 2e-i,ii: Chi-squared tests comparing clustering results
```{r}
cat("\n\n=== Chi-squared tests comparing clustering results ===\n\n")

spectral_best_assignments <- purrr::map(spectral_all_genes, "best_clusters")

cat("--- Pairwise comparisons between different gene counts ---\n")
gene_count_names <- names(spectral_best_assignments)
gene_pairs <- combn(gene_count_names, 2, simplify = FALSE)

safe_chisq <- function(a, b) {
  if (length(a) != length(b)) return(list(ok=FALSE, reason="length_mismatch", test=NULL))
  if (length(a) < 2L)       return(list(ok=FALSE, reason="too_few_samples", test=NULL))
  tab <- table(a, b)
  if (nrow(tab) < 2L || ncol(tab) < 2L)
    return(list(ok=FALSE, reason="table_not_2x2", test=NULL))
  list(ok=TRUE, reason=NA_character_, test=suppressWarnings(chisq.test(tab)))
}

chi_sq_gene_pairs <- purrr::map_dfr(gene_pairs, function(pair) {
  name1 <- pair[1]; name2 <- pair[2]
  g1 <- spectral_all_genes[[name1]]; g2 <- spectral_all_genes[[name2]]

  clust1 <- spectral_best_assignments[[name1]]
  clust2 <- spectral_best_assignments[[name2]]

  
  samples <- intersect(names(clust1), names(clust2))
  clust1 <- clust1[samples]
  clust2 <- clust2[samples]

  res <- safe_chisq(clust1, clust2)

  if (!res$ok) {
    message(sprintf("[skip %s vs %s] %s | n=%d", name1, name2, res$reason, length(samples)))
    return(tibble::tibble(
      comparison = "gene_counts",
      gene_count_a = g1$gene_count,
      gene_count_b = g2$gene_count,
      k_a = g1$best_k,
      k_b = g2$best_k,
      chi_sq_statistic = NA_real_,
      df = NA_real_,
      p_value = NA_real_,
      note = res$reason
    ))
  }

  tibble::tibble(
    comparison = "gene_counts",
    gene_count_a = g1$gene_count,
    gene_count_b = g2$gene_count,
    k_a = g1$best_k,
    k_b = g2$best_k,
    chi_sq_statistic = unname(res$test$statistic),
    df = unname(res$test$parameter),
    p_value = res$test$p.value,
    note = NA_character_
  )
})

print(chi_sq_gene_pairs)
```
```{r}
suppressPackageStartupMessages({
  library(pheatmap)
  library(RColorBrewer)
  library(glue)
})

# --- data prep (same as before) ---
heatmap_matrix <- log2(expression_top5000 + 1)
heatmap_matrix <- t(scale(t(heatmap_matrix)))
heatmap_matrix[!is.finite(heatmap_matrix)] <- 0
heatmap_samples <- colnames(heatmap_matrix)

#align annotations
spectral_col <- as.character(spectral_clusters_5000_best[heatmap_samples])
exposure_col <- as.character(exposure_lookup[heatmap_samples])
exposure_col[is.na(exposure_col)] <- "Unknown"

#factors AFTER alignment
spectral_factor <- factor(spectral_col)
exposure_factor <- factor(exposure_col)

# annotation data frame 
annotation_col <- data.frame(
  Exposure = exposure_factor,
  Spectral = spectral_factor,
  row.names = heatmap_samples
)

# annotation colors
mk_cols <- function(n, pal) {
  colorRampPalette(brewer.pal(8, pal))(max(1L, n))
}
ann_colors <- list(
  Exposure = setNames(mk_cols(nlevels(exposure_factor), "Dark2"), levels(exposure_factor)),
  Spectral = setNames(mk_cols(nlevels(spectral_factor), "Set3"),  levels(spectral_factor))
)

# heatmap colors 
hm_cols <- colorRampPalette(c("navy", "white", "firebrick3"))(101)

title_text <- if (exists("best_k_5000") && !is.na(best_k_5000)) {
  glue("Top 5000 genes - Spectral (k = {best_k_5000})")
} else "Top 5000 genes - Spectral"


if (!exists("plots_dir")) plots_dir <- "plots"
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

dir.create("plots", showWarnings = FALSE, recursive = TRUE)

ph <- pheatmap(
  mat = heatmap_matrix,
  color = hm_cols,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  show_colnames = FALSE,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  main = title_text,
  legend = TRUE,
  fontsize = 10,
  border_color = NA,
  filename = file.path("plots", "spectral_heatmap_top5000.png")  # writes to disk
)

print(normalizePath(file.path("plots", "spectral_heatmap_top5000.png"), winslash = "/"))

```

### 4a: Statistical comparison with Assignment 1 groups

```{r spectral-assignment1-comparison, cache=TRUE, cache.lazy=FALSE}
cat("\n\n=== Comparing clusters with Assignment 1 groups ===\n\n")

chi_sq_vs_assignment1 <- map_dfr(names(spectral_all_genes), function(name) {
  gene_n <- spectral_all_genes[[name]]$gene_count
  best_k <- spectral_all_genes[[name]]$best_k
  clusters <- spectral_all_genes[[name]]$best_clusters
  
  
  samples <- names(clusters)
  exposures <- exposure_lookup[samples]
  exposures[is.na(exposures)] <- "Unknown"
  
  
  tab <- table(
    exposure = factor(exposures),
    cluster = clusters
  )
  test <- suppressWarnings(chisq.test(tab))
  
  tibble(
    comparison = "clusters_vs_assignment1",
    gene_count = gene_n,
    k = best_k,
    chi_sq_statistic = unname(test$statistic),
    df = unname(test$parameter),
    p_value = test$p.value
  )
})

cat("--- Chi-squared tests: Spectral clusters vs Assignment 1 groups ---\n")
print(chi_sq_vs_assignment1)

readr::write_tsv(k_comparison_5000, file.path(results_dir, "spectral_k_comparison_5000.tsv"))
save_table_png(k_comparison_5000, file.path(plots_dir, "spectral_k_comparison_5000.png"))


save_table_png(chi_sq_k_comparison, file.path(plots_dir, "spectral_chi_sq_k_comparison_5000.png"))


install.packages("gridExtra")
library(gridExtra); library(grid)

save_table_png <- function(df, path, base_size = 10, dpi = 200, min_w = 8, min_h = 4){
  tb <- tableGrob(df, rows = NULL, theme = ttheme_minimal(base_size = base_size))
  png(tempfile(fileext = ".png"), width = 10, height = 10, units = "in", res = dpi)
  grid.newpage(); grid.draw(tb)
  w_in <- convertWidth(sum(tb$widths), "in", valueOnly = TRUE)
  h_in <- convertHeight(sum(tb$heights), "in", valueOnly = TRUE)
  dev.off()
  w <- max(min_w, w_in); h <- max(min_h, h_in)
  png(path, width = w, height = h, units = "in", res = dpi)
  grid.newpage(); grid.draw(tb)
  dev.off()
}

save_table_png(k_comparison_5000, file.path("plots","spectral_k_comparison_5000.png"))
save_table_png(chi_sq_k_comparison, file.path("plots","spectral_chi_sq_k_comparison_5000.png"))
save_table_png(best_k_summary, file.path("plots","spectral_best_k_summary.png"))
save_table_png(chi_sq_gene_pairs, file.path("plots","spectral_chi_sq_gene_pairs.png"))
save_table_png(chi_sq_vs_assignment1, file.path("plots","spectral_chi_sq_vs_assignment1.png"))
save_table_png(all_chi_sq_tests, file.path("plots","spectral_all_chi_sq_tests.png"))
```

### 4b: Adjust for multiple hypothesis testing

```{r spectral-adjust-pvalues, cache=TRUE}
cat("\n\n=== Adjusting p-values for multiple testing ===\n\n")
library(dplyr)
#combine all chi-squared tests
all_chi_sq_tests <- bind_rows(
  chi_sq_gene_pairs,
  chi_sq_vs_assignment1 %>% 
    transmute(
      comparison,
      gene_count_a = gene_count,
      gene_count_b = gene_count,
      k_a = k,
      k_b = k,
      chi_sq_statistic,
      df,
      p_value
    )
) %>%
  mutate(p_adj_BH = p.adjust(p_value, method = "BH")) %>%
  arrange(comparison, gene_count_a, gene_count_b)

cat("--- All chi-squared tests with adjusted p-values ---\n")
print(all_chi_sq_tests)

if (!exists("results_dir")) {
  results_dir <- "results"
}
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

readr::write_tsv(best_k_summary, file.path(results_dir, "spectral_best_k_summary.tsv"))
readr::write_tsv(all_chi_sq_tests, file.path(results_dir, "spectral_chisq_tests.tsv"))
readr::write_tsv(k_comparison_5000, file.path(results_dir, "spectral_k_comparison_5000genes.tsv"))


assign("spectral_clusters_5000", 
       tibble(sample_id = names(spectral_clusters_5000_best), 
              cluster = as.character(spectral_clusters_5000_best)), 
       envir = .GlobalEnv)
assign("spectral_best_k", best_k_5000, envir = .GlobalEnv)
```

### 5: Summary

**How changing k affects results (5000 genes):** I applied spectral clustering using the top 5000 most variable genes from the normalized expression matrix. The algorithm was run for k values ranging from 2 to 8, and I computed the average silhouette score for each k to assess clustering quality. The highest silhouette score occurred at k = 2, indicating that a two-cluster solution best separates the samples in this dataset. Increasing k consistently lowered the silhouette, suggesting that additional clusters do not represent meaningful substructure. 

**Effect of gene count on clustering:** I statistically compared cluster memberships between all pairs of k values (2–8) using chi-squared tests of independence. This tested whether the partitions at different k values produce consistent group assignments. All pairwise chi-squared tests between k values were highly significant (p < 0.001 for most), with large chi-squared statistics. The strong significance indicates that changing k meaningfully changes the cluster composition, not just by random variation. In other words, each increment of k subdivides the two primary groups rather than revealing new distinct patterns — confirming that k = 2 captures the major biological separation.



**Heatmap with annotations:** I created a heatmap using pheatmap() of the top 5,000 scaled genes, annotated by spectral clusters (k = 2 ) and exposure group. The heatmap clearly shows two broad expression patterns across samples that align with the two spectral clusters. Visual comparison of the annotation bars reveals a partial overlap between Spectral clusters and exposure status: Cluster 1 contains many DENV samples. Cluster 2 includes a higher proportion of naive samples. This visual concordance supports the idea that the spectral algorithm partially recovered the biological grouping structure encoded in exposure metadata.

**Statistical validation:** I repeated spectral clustering for 10, 100, 1,000, 5,000, and 10,000 most variable genes. For each, the best k (max silhouette) was selected, and results were compared via chi-squared tests.
cross most gene subsets, k = 2 remained optimal, showing that the global two-group structure is stable. At 100 genes, the algorithm briefly favored k = 6, suggesting overfitting or noise sensitivity at very small gene subsets.
 Silhouette scores decreased with extremely small (10) or very large (10,000) gene sets — consistent with too little or too much noise diluting signal.
This implies that using an intermediate set (1,000–5,000 genes) best balances stability and biological signal for spectral clustering.
```

